---
title: "AI-powered bioinformatic analysis using R, #8"
subtitle: "Strings, Iteration & Functional Programming"
author: "Wei-Hua Chen (CC BY-NC 4.0)"
institute: "HUST, China"
date: today

CJKmainfont: "Arial"

format: 
  beamer:
    aspectratio: 1610
    navigation: vertical
    theme: AnnArbor
    colortheme: beaver
    fonttheme: structurebold
    highlight: tango
    slide_level: 2
    toc: true
    pdf-engine: xelatex
    header-includes:
      - \AtBeginSubsection{}

execute:
  echo: true
  fig-width: 8
  fig-height: 4.5
---

# String Basics

## What Are Strings?

Strings are sequences of characters. In bioinformatics, strings are everywhere:

- DNA/RNA sequences: `"ATCGGCGCG..."`
- Gene names: `"TP53"`, `"BRCA1"`
- File paths, sample IDs, annotations

## Some exmaples

\footnotesize
```{r}
library(tidyverse)
library(stringr)

string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'

# Multiline string
(string3 <- "a multiline
string")
writeLines(string3)  # see the actual content
```
\normalsize

## Special Characters

\footnotesize
```{r}
# Double and single quotes
(double_quote <- "\"")
(single_quote <- '\'')

# Unicode
(x <- "\u00b5")  # µ (micro sign)

# Backslash — note the difference!
(y <- "\\")
writeLines(y)
```
\normalsize

## String Length

\footnotesize
```{r}
# Base R
nchar(c("a", "R for data science", NA))

# stringr
str_length(c("a", "R for data science", NA))
```
\normalsize

## String Concatenation

\footnotesize
```{r}
# Base R — paste()
paste("a", "b", "c", sep = "")

# stringr — str_c()
str_c("a", "b", "c")

# Vectorized
paste(c("a", "b", "c"), 2, sep = "")
str_c(c("a", "b", "c"), 2)
```
\normalsize

## String Comparison

\footnotesize
```{r}
# Direct comparison (lexicographic)
"A" > "abc"

library(pracma)
strcmp("chen", "chenweihua")    # case-sensitive
strcmpi("chen", "CHEN")         # case-insensitive
```
\normalsize

## Case Conversion & Other Basics

\footnotesize
```{r}
toupper(letters[1:10])
tolower(LETTERS[1:5])

library(stringi)
stri_reverse("ABC")
```
\normalsize

**Tip**: `stringi` functions start with `stri_`, `stringr` functions start with `str_`.

# Regular Expressions (Regex)

## What is Regex?

**A sequence of characters that define a search pattern.**

\footnotesize
```{r}
# [ab] means "a or b"
c("abc", "chen", "liu", "blah") %>% str_subset("[ab]")

# \d+ means "one or more digits"
c("a1334bc", "ch13e_45n", "liu", "bl00ah") %>% str_extract("\\d+")
```
\normalsize

**Useful tools**: [regex101.com](https://regex101.com), [regexr.com](https://regexr.com)

## Four Tasks of Regex

1. **Match patterns** — what characters?
2. **Match rules** — which character classes?
3. **Position rules** — where in the string?
4. **Quantity rules** — how many repetitions?

## 1. Character Classes & Groups

| Pattern | Meaning |
|---------|---------|
| `.` | any character |
| `[abc]` | a, b, or c |
| `[^abc]` | anything except a, b, c |
| `[a-z]` | any lowercase letter |
| `[A-Z]` | any uppercase letter |
| `[0-9]` | any digit |
| `(ab\|cd)` | "ab" or "cd" |

## 1. Special Character Classes

| Shorthand | Meaning |
|-----------|---------|
| `\d` | digit `[0-9]` |
| `\D` | non-digit |
| `\w` | word character `[A-Za-z0-9_]` |
| `\W` | non-word |
| `\s` | whitespace (space, tab, newline) |
| `\S` | non-whitespace |

\footnotesize
```{r}
"abc_123_??$$^" %>% str_detect("\\s+")   # has whitespace?
"abc_123_??$$^" %>% str_detect("\\d+")   # has digits?
"abc_123_??$$^" %>% str_detect("\\w+")   # has word chars?
```
\normalsize

## 2. Position Rules — Anchors

| Anchor | Meaning |
|--------|---------|
| `^` | start of string |
| `$` | end of string |
| `\b` | word boundary |

\footnotesize
```{r}
# Ends with "wei"
c("chen wei hua", "chen wei", "chen") %>% str_subset("wei$")

# Word ending in "wei"
c("chen wei hua", "chen wei", "chen") %>% str_subset("wei\\b")
```
\normalsize

## 3. Quantity Rules — Quantifiers

| Quantifier | Meaning |
|------------|---------|
| `+` | 1 or more |
| `*` | 0 or more |
| `?` | 0 or 1 |
| `{n}` | exactly n |
| `{n,}` | n or more |
| `{n,m}` | between n and m |

\footnotesize
```{r}
"1234abc" %>% str_extract("\\d+")       # all digits
"1234abc" %>% str_extract("\\d{3}")     # exactly 3 digits
"1234abc" %>% str_extract("\\d{5,6}")   # 5-6 digits → NA
"1234abc" %>% str_extract("\\d{2,6}")   # 2-6 digits
```
\normalsize

# Regex Tasks in stringr

## 1. Detect Patterns

\footnotesize
```{r}
# Base R
grep("\\d+", c("123", "abc", "wei555hua"))
grepl("\\d+", c("123", "abc", "wei555hua"))

# stringr
c("123", "abc", "wei555hua") %>% str_detect("\\d+")
```
\normalsize

## 1. Count Matches

\footnotesize
```{r}
x <- c("why", "video", "cross", "extra", "deal", "authority")
str_detect(x, "[aeiou]")    # contains vowel?
str_count(x, "[aeiou]")     # how many vowels?
```
\normalsize

## 2. Locate Patterns

\footnotesize
```{r}
# Base R
regexpr("\\d+", c("123", "abc", "wei555hua"))

# stringr
c("123", "abc", "wei555hua") %>% str_locate("\\d+")
```
\normalsize

## 3. Extract Patterns

\footnotesize
```{r}
c("123", "abc", "wei555hua") %>% str_extract("\\d+")
c("123", "abc", "wei555hua") %>% str_match("\\d+")
```
\normalsize

## str_extract vs. str_match

\footnotesize
```{r}
x
str_extract(x, "(.)[aeiou](.)")            # extract match
str_match(x, "(.)[aeiou](.)")        # extract groups around vowel
```
\normalsize

`str_match()` returns capture groups `(...)` as separate columns!

## Extract All Occurrences

\footnotesize
```{r}
x
str_extract_all(x, "[aeiou]+")
str_match_all(x, "[aeiou]+")
```
\normalsize

## 4. Replace Patterns

\footnotesize
```{r}
str_replace(c("123", "abc", "wei555hua"), "\\d+", "###")
str_replace_all("123_abc_456_789", "\\d+", "###")
```
\normalsize

## 5. Split by Pattern

\footnotesize
```{r}
str_split(x, "")
```
\normalsize

## Advanced Example: Chromosome Coordinates

\footnotesize
```{r}
dat <- tibble(chrom = readLines(textConnection(
  "chr11:69464719-69502928
chr7:55075808-55093954
chr8:128739772-128762863
chr3:169389459-169490555
chr17:37848534-37877201
chr19:30306758-30316875
chr1:150496857-150678056
chr12:69183279-69260755
chr11:77610143-77641464
chr8:38191804-38260814
chr12:58135797-58156509")))

# Split into chr, start, end
dat$chrom %>% str_split('[:-]', simplify = TRUE)
```
\normalsize

## Alternative: tidyr::extract()

\footnotesize
```{r}
library(tidyr)
extract(dat, chrom, into = c('chr', 'chrStart', 'chrEnd'),
        '([^:]+):([^-]+)-(.*)', convert = TRUE)
```
\normalsize

# Iteration: for Loops

## for Loop Basics

\footnotesize
```{r message=FALSE, warning=FALSE}
library(tidyverse)

df <- tibble(a = rnorm(100), b = rnorm(100), c = rnorm(100), d = rnorm(100))

# Compute row means — pre-allocate output!
res1 <- vector("double", nrow(df))
for (row_idx in 1:nrow(df)) {
  res1[row_idx] <- mean(as.numeric(df[row_idx, ]))
}

# Compute column means
res2 <- vector("double", ncol(df))
for (col_idx in 1:ncol(df)) {
  res2[col_idx] <- mean(df[[col_idx]])
}
```
\normalsize

**Golden rule**: Always **pre-allocate** the output vector! `vector("double", N)`

## Vectorized Alternatives to for Loops

\footnotesize
```{r}
# Much faster than for loops
rowMeans(df)
colMeans(df)

# Also available:
# rowSums(), colSums()
```
\normalsize

# The apply Family

## apply() — Row/Column Operations

`apply(X, MARGIN, FUN, ...)`

- **MARGIN = 1**: rows
- **MARGIN = 2**: columns

\footnotesize
```{r}
df %>% apply(1, median)    # row medians
df %>% apply(2, median)    # column medians
```
\normalsize

## apply() with Custom Functions

\footnotesize
```{r}
df %>% apply(2, function(x) {
  return(c(n = length(x), mean = mean(x), median = median(x)))
})
```
\normalsize

**Note**: Most row operations can be replaced by `dplyr`.

## tapply() — Grouped Operations

`tapply(X, INDEX, FUN)`

Group **X** by **INDEX**, apply **FUN** to each group.

\footnotesize
```{r warning=FALSE, message=FALSE}
library(magrittr)
# mpg by cylinder
mtcars %$% tapply(mpg, cyl, mean)
```
\normalsize

## tapply vs. dplyr

\footnotesize
```{r warning=FALSE, message=FALSE}
# tapply
mtcars %$% tapply(mpg, cyl, mean)

# dplyr — clearer, more readable
mtcars %>% group_by(cyl) %>% summarise(mean = mean(mpg))
```
\normalsize

**Both tapply and dplyr operate on rows!**

## lapply() and sapply()

Column/list operations — iterate over each element:

- **vector**: one element at a time
- **data.frame / tibble / matrix**: one column at a time
- **list**: one member at a time

\footnotesize
```{r}
df %>% lapply(mean)    # returns list
```
\normalsize

## sapply() — Type-stable

\footnotesize
```{r}
df %>% sapply(mean)    # returns vector
```
\normalsize

## lapply/sapply with Custom Functions

\footnotesize
```{r}
list(a = 1:10, b = letters[1:5], c = LETTERS[1:8]) %>%
  sapply(function(x) { length(x) })
```
\normalsize

**Key**: `lapply`/`sapply` = column-wise operations, equivalent to `apply(x, 2, FUN)`.

# purrr: Modern Iteration

## What is purrr?

Part of the **tidyverse**, provides consistent, type-stable iteration functions.

`map()` = `lapply()` replacement:

- Iterates over each column (tibble) or slot (list)
- Applies FUN
- Returns results as a list

## map() — Column/Slot Operations

\footnotesize
```{r}
df %>% map(summary)
```
\normalsize

## Typed map Functions (sapply replacement)

| Function | Returns |
|----------|---------|
| `map_lgl()` | logical vector |
| `map_int()` | integer vector |
| `map_dbl()` | double vector |
| `map_chr()` | character vector |
| `map_df()` | data frame (row-binding) |

\footnotesize
```{r}
df %>% map_dbl(mean)
```
\normalsize

## Advanced map: split + cor.test

\footnotesize
```{r fig.height=4, fig.width=10}
# For each cylinder group, correlate mpg and wt
mtcars %>%
  split(.$cyl) %>%
  map(~ cor.test(.$wt, .$mpg)) %>%
  map_dbl(~ .$estimate)
```
\normalsize

The code logic:

- split the data frame by `cyl`
- for each group, run `cor.test(wt, mpg)`
- extract the `estimate` from the result

## The code above is equivalent to:

\footnotesize
```{r}
# For each cylinder group, correlate mpg and wt
mtcars %>%
  group_by(cyl) %>%
  summarise(cor = cor.test(wt, mpg)$estimate)
```

## The ~ (tilde) Syntax

Short for `function(x)`:

```{r eval=FALSE}
# Full form
map(function(df) { cor.test(df$wt, df$mpg) })

# Shorthand — ~ replaces function, . replaces the argument
map(~ cor.test(.$wt, .$mpg))
```

## map2() — Two-Input Mapping

\footnotesize
```{r}
mu <- list(5, 10, -3)
sigma <- list(1, 5, 10)
map2(mu, sigma, rnorm, n = 5) %>% str()
```
\normalsize

What is the output?

- generate 5 random numbers for each pair of `mu` (as mean) and `sigma` (as standard deviation)

## pmap() — Multi-Input Mapping

\footnotesize
```{r}
params <- tribble(
  ~mean, ~sd, ~n,
     5,    1,  1,
    10,    5,  3,
    -3,   10,  5
)
params %>% pmap(rnorm)
```
\normalsize

## reduce() — Combine Elements

\footnotesize
```{r}
# Full join multiple tables
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)
dfs %>% reduce(full_join)
```
\normalsize

## reduce() for Set Operations

\footnotesize
```{r}
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)
vs %>% reduce(intersect)    # common elements
```
\normalsize

## accumulate() — See Intermediate Steps

\footnotesize
```{r}
(x <- sample(10))
x %>% accumulate(`+`)    # cumulative sum, step by step
```
\normalsize

## keep() and discard()

\footnotesize
```{r}
# Keep factor columns
iris %>% keep(is.factor)
```

## discard()

\footnotesize
```{r}
# Discard factor columns
iris %>% discard(is.factor)
```

\normalsize

# Writing Custom Functions

## Function Syntax

\footnotesize

```{r}
# Basic structure
my_function <- function(arg1, arg2 = default_value) {
  # body
  return(result)
}

# Example
calc_cv <- function(x) {

    ## if no return value is specified, the last expression is returned
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100
}

calc_cv(mtcars$mpg)
```

## Defensive Programming

Always validate inputs:

\footnotesize
```{r}
safe_mean <- function(x) {
  stopifnot(is.numeric(x))         # validate input
  if (length(x) == 0) return(NA)   # handle edge case
  mean(x, na.rm = TRUE)
}

# safe_mean(c("a", "b"))  # would error with clear message
safe_mean(c(1, 2, 3, NA))
```
\normalsize

# Parallel Computing (Brief)

## Why Parallel Computing?

- Modern computers have multiple CPU cores
- Many bioinformatics tasks are "embarrassingly parallel"
- Speed up by running tasks simultaneously on multiple cores

## Three Steps of Parallel Computing

1. **Split** tasks and distribute to workers
2. **Compute** independently on each worker
3. **Collect** and combine results

## Key Packages

- **`parallel`**: detect CPU cores, `parLapply()`
- **`doParallel`**: register parallel backend
- **`foreach`**: provides `%do%` (sequential) and `%dopar%` (parallel)
- **`iterators`**: split data for parallel tasks

## Simple Example

\footnotesize
```{r}
library(doParallel)
library(foreach)
library(iterators)

# Detect CPUs
(cpus <- parallel::detectCores())

# Register parallel backend
registerDoParallel(cpus - 1)   # leave 1 core free

# Parallel loop
data(iris)
x <- iris[which(iris[, 5] != "setosa"), c(1, 5)]
trials <- 10000

r <- foreach(icount(trials), .combine = rbind) %dopar% {
  ind <- sample(100, 100, replace = TRUE)
  result1 <- glm(x[ind, 2] ~ x[ind, 1], family = binomial(logit))
  coefficients(result1)
}
```
\normalsize

## Sequential vs. Parallel: Time Comparison

Let's compare running time of sequential vs. parallel for a realistic bioinformatics task — performing many t-tests on simulated gene expression data:

\footnotesize
```{r}
library(doParallel)
library(foreach)

# Simulate: 5000 genes × 20 samples (10 case, 10 control)
set.seed(123)
n_genes <- 5000
expr_data <- matrix(rnorm(n_genes * 20, mean = 10, sd = 2), nrow = n_genes)
colnames(expr_data) <- c(paste0("Case_", 1:10), paste0("Ctrl_", 1:10))
group <- rep(c("Case", "Ctrl"), each = 10)

# --- Sequential ---
time_seq <- system.time({
  res_seq <- foreach(i = 1:n_genes, .combine = c) %do% {
    t.test(expr_data[i, group == "Case"], 
           expr_data[i, group == "Ctrl"])$p.value
  }
})
```

## Sequential vs. Parallel: Time Comparison, cont.

\footnotesize
```{r}
# Register parallel backend
registerDoParallel(parallel::detectCores() - 1)

# --- Parallel ---
time_par <- system.time({
  res_par <- foreach(i = 1:n_genes, .combine = c) %dopar% {
    t.test(expr_data[i, group == "Case"], 
           expr_data[i, group == "Ctrl"])$p.value
  }
})

stopImplicitCluster()
```

## Sequential vs. Parallel: Time Comparison, cont.

\footnotesize
```{r}
# Compare timing
rbind(Sequential = time_seq[1:3], Parallel = time_par[1:3])

# Speedup
cat(sprintf("\nSpeedup: %.1fx faster with parallel computing\n",
            time_seq["elapsed"] / time_par["elapsed"]))
```
\normalsize

**Key takeaway**: For independent, repetitive computations (e.g., per-gene statistics, bootstrapping, permutation tests), parallel computing can dramatically reduce runtime.

## foreach Parameters

- **`.combine`**: how to combine results (`'c'`, `'rbind'`, `'cbind'`, or a custom function)
- **`.packages`**: packages to load on each worker (e.g., `c("tidyverse")`)

\footnotesize
```{r eval=FALSE}
# Example with .packages
res <- foreach(df = iter(mtcars %>% split(.$cyl)), 
               .combine = 'rbind',
               .packages = c("ggplot2")) %dopar% {
  p <- ggplot(df, aes(x = wt, y = mpg)) + geom_point()
  return(p)
}

# ALWAYS close the cluster!
stopImplicitCluster()
```
\normalsize

## furrr: purrr + future

A simpler approach — combine `purrr` with `future`:

\footnotesize
```{r eval=FALSE}
library(furrr)
plan(multisession, workers = 4)  # use 4 cores

# Replace map() with future_map()
result <- mtcars %>%
  split(.$cyl) %>%
  future_map(~ cor.test(.$wt, .$mpg))
```
\normalsize

# Summary

## What We Covered

| Topic | Key Functions |
|-------|---------------|
| **String basics** | `str_length`, `str_c`, `toupper`, `tolower` |
| **Regex** | `str_detect`, `str_count`, `str_locate`, `str_extract`, `str_match`, `str_replace`, `str_split` |
| **for loops** | pre-allocation, `vector("double", N)` |
| **apply family** | `apply()`, `tapply()`, `lapply()`, `sapply()` |
| **purrr** | `map()`, `map_dbl()`, `map2()`, `pmap()`, `reduce()`, `accumulate()`, `keep()`, `discard()` |
| **Custom functions** | `function()`, `stopifnot()`, input validation |
| **Parallel computing** | `foreach` + `%dopar%`, `furrr::future_map()` |

## Next Talk

**Talk 09**: Statistical Analysis & Machine Learning Basics

## Homework

Complete `Homework_AI_powered_R/talk08-homework.qmd`
