---
title: "AI-powered bioinformatic analysis using R, #3"
subtitle: "R Language Basics II — Factors, Lists, Data Frames & File I/O"
author: "Wei-Hua Chen (CC BY-NC 4.0)"
institute: "HUST, China"
date: today

CJKmainfont: "Arial"

format: 
  beamer:
    aspectratio: 1610
    navigation: vertical
    theme: AnnArbor
    colortheme: beaver
    fonttheme: structurebold
    highlight: tango
    slide_level: 2
    toc: true
    pdf-engine: xelatex
    header-includes:
      - \AtBeginSection{}
      - \AtBeginSubsection{}

execute:
  echo: true
  fig-width: 8
  fig-height: 4.5
---

# Factors

## What is a Factor?

In bioinformatics, many variables are **categorical**:
- Treatment groups: "Control", "Drug A", "Drug B"
- Tissue types: "Liver", "Kidney", "Brain"
- Genotypes: "AA", "Aa", "aa"
- Disease status: "Healthy", "Mild", "Severe"

**Factors** are R's way of handling categorical data. They store values as **integers** with a set of **labels** (levels).

:::: {.callout-tip}
Factors look like characters but behave like **coded integers** — this is crucial for statistical modeling!
::::

## Creating Factors

Use `factor()` to create a factor from a vector:

\footnotesize

```{r}
# A character vector of treatment groups
treatment <- c("Control", "Drug", "Control", "Drug", "Control")

# Convert to factor
( treat_f <- factor(treatment) )
```

## Creating Factors, cont.

\footnotesize

Check the internal structure:

```{r}
class(treat_f)
str(treat_f)       # shows integer codes + levels
levels(treat_f)    # unique categories
```

## Factor Levels — Default Order

By default, levels are ordered **alphabetically**:

\footnotesize

```{r}
group <- c("High", "Low", "Medium", "Low", "High")
factor(group)
```

\vspace{0.3cm}

\footnotesize

```{r}
# Levels are: High, Low, Medium (alphabetical)
# This may NOT be what you want for plotting or modeling!
```

\normalsize

For plots and statistical models, you often need a **meaningful order**:
`Low < Medium < High` (not alphabetical).

## Controlling Level Order

Specify the `levels` argument explicitly:

\footnotesize

```{r}
( group_f <- factor(group, 
                    levels = c("Low", "Medium", "High")) )
```

\vspace{0.3cm}

```{r}
# Now levels are in the correct logical order
levels(group_f)
```

## Ordered Factors

Use `ordered = TRUE` (or `ordered()`) to create an **ordered factor**:

\footnotesize

```{r}
( severity <- ordered(c("Mild", "Severe", "Moderate", "Mild"),
                       levels = c("Mild", "Moderate", "Severe")) )
```

\vspace{0.3cm}

```{r}
# Comparisons work on ordered factors
severity[1] < severity[2]    # "Mild" < "Severe" → TRUE
```

\normalsize

:::: {.callout-warning}
Only use `ordered = TRUE` when the categories have a **genuine** order (e.g., disease stage). Don't order nominal categories like "Red", "Green", "Blue".
::::

## Factor Manipulation with `forcats`

The **forcats** package (part of tidyverse) provides tools for factor management:

\footnotesize

```{r}
library(forcats)  # or library(tidyverse)

# fct_relevel()  — reorder levels manually
fct_relevel(treat_f, "Drug", "Control")

# fct_rev()      — reverse level order
fct_rev(treat_f)

# fct_infreq()   — order by frequency
fct_infreq(treat_f)
```

## `forcats` — Renaming & Combining Levels

\footnotesize

```{r}
x <- factor(c("apple", "bear", "banana", "dates", "bear"))

# fct_recode()  — rename levels
fct_recode(x, fruit = "apple", fruit = "banana", 
              fruit = "dates", animal = "bear")

# fct_lump()    — lump small groups together
y <- factor(letters[rpois(100, 5)])
fct_lump(y, n = 5)    # keep top 5, lump rest as "Other"
```

## Factors in Bioinformatics — Example

\footnotesize

```{r}
# Gene expression categories in a cancer study
expr_level <- c("Under-expressed", "Normal", "Over-expressed",
                "Over-expressed", "Normal", "Under-expressed",
                "Normal", "Normal", "Over-expressed", "Under-expressed")

( expr_f <- factor(expr_level, 
                    levels = c("Under-expressed", "Normal", "Over-expressed"),
                    ordered = TRUE) )

# Count how many genes in each category
table(expr_f)

# Summary statistics by category
summary(expr_f)
```

## Why Factors Matter

1. **Statistical models**: `lm()` and `glm()` automatically create dummy variables from factors
2. **Plot ordering**: `ggplot2` respects factor level order on axes
3. **Memory efficiency**: factors store integers internally, not full strings
4. **Data validation**: only values in `levels` are allowed

\vspace{0.3cm}

:::: {.callout-tip}
**AI tip**: Ask CodeBuddy: "My ggplot bar chart shows categories in wrong order — how do I reorder using `fct_relevel()`?"
::::

# Lists

## What is a List?

A **list** is a container that can hold **different types** of objects — unlike vectors which require uniform types. Lists are R's most flexible data structure.

\footnotesize

```{r}
# A list can contain anything
my_list <- list(
  name       = "Gene TP53",
  chromosome = 17,
  is_oncogene = TRUE,
  expression = c(2.3, 4.5, 1.8, 3.2, 5.1),
  metadata   = list(study = "TCGA", cancer = "BRCA")
)
my_list
```

## List Structure

Lists are **recursive** — they can contain other lists:

\footnotesize

```{r}
str(my_list)       # inspect the structure
length(my_list)    # number of top-level elements
names(my_list)     # element names
```

## List Structure, cont.

\footnotesize

Lists are everywhere in bioinformatics R:
- Statistical model outputs (`lm()`, `glm()`)
- Genomic ranges (Bioconductor `GRanges`)
- Seurat single-cell objects
- JSON/API responses

## Creating Lists

\footnotesize

```{r}
# Unnamed list
list(1:5, letters[1:3], TRUE)

# Named list (preferred for readability)
list(numbers = 1:5, chars = letters[1:3], flag = TRUE)

# Empty list
list()
```

## Creating Lists, cont.

\footnotesize

```{r}
# List from vectors of different lengths
list(
  patient_ids  = c("P001", "P002", "P003"),
  ages         = c(45, 62, 38),
  diagnoses    = c("Healthy", "T2D", "Healthy"),
  medications  = c("None", "Metformin"),       # shorter — no recycling!
  lab_results  = data.frame(glucose = c(90, 140, 95),
                            insulin = c(5.2, 12.1, 4.8))
)
```

## Accessing List Elements — `$` and `[[ ]]`

The `$` operator extracts a single named element:

\footnotesize

```{r}
my_list$name          # returns the element itself
my_list$expression    # returns the numeric vector
```

\vspace{0.3cm}

`[[ ]]` extracts by position or name:

```{r}
my_list[[1]]          # first element
my_list[["name"]]     # by name (equivalent to $)
```

## Accessing List Elements — `[ ]` vs `[[ ]]`

**Single bracket `[ ]`** returns a **sub-list**:

\footnotesize

```{r}
my_list[1]            # a list containing the first element
class(my_list[1])     # "list"
```

## Accessing List Elements — `[ ]` vs `[[ ]]`, cont.

\footnotesize

**Double bracket `[[ ]]`** returns the **element itself**:

```{r}
my_list[[1]]          # the actual value
class(my_list[[1]])   # "character"
```

\normalsize

:::: {.callout-tip}
Think of it like a train: `[ ]` gives you a train car (still a train), `[[ ]]` gives you what's inside the car.
::::

## Deep Access in Nested Lists

\footnotesize

```{r}
# Accessing nested elements
my_list$metadata$study        # "TCGA"
my_list[["metadata"]][["cancer"]]  # "BRCA"

# Equivalent chaining
my_list$metadata[[2]]         # also "BRCA" (2nd element of metadata)
```

## Modifying Lists

\footnotesize

```{r}
# Add a new element
my_list$gene_length <- 19149  # base pairs

# Modify an existing element
my_list$expression <- c(3.1, 4.8, 2.0, 3.5, 5.5)

# Remove an element
my_list$metadata <- NULL

str(my_list)
```

\normalsize

Adding an element to a list is as simple as assigning to a new name with `$`.

## Lists in Bioinformatics — Practical Example

\footnotesize

```{r}
# Simulate a differential expression result
de_result <- list(
  gene = "BRCA1",
  log2FC = 2.34,
  p_value = 0.00042,
  fdr = 0.015,
  samples = c("Tumor1", "Tumor2", "Normal1", "Normal2")
)

# Extract p-value for downstream filtering
de_result$p_value

# Check if gene is significant (FDR < 0.05)
de_result$fdr < 0.05
```

# Data Frames & Tibbles

## What is a Data Frame?

A **data frame** is a 2-dimensional table where each column can have a different type:

\footnotesize

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Built-in example dataset
library(tidyverse)
library(kableExtra)
kbl(head(mpg), booktabs = TRUE)
```

\vspace{0.3cm}

\normalsize

- Each **column** is a vector of a single type
- Each **row** is one observation
- The most fundamental data structure for tabular data in R

## Data Frame Properties

\footnotesize

```{r}
nrow(mpg)      # number of rows
ncol(mpg)      # number of columns
dim(mpg)       # both
str(mpg)       # structure overview
```

## Data Frame Properties, cont.

\footnotesize

```{r}
# Quick peek
head(mpg, n = 3)     # first 3 rows
tail(mpg, n = 3)     # last 3 rows
```

## Creating Data Frames

\footnotesize

```{r}
# data.frame() function
( dat <- data.frame(
  gene   = c("BRCA1", "TP53", "EGFR", "MYC"),
  logFC  = c(2.3, -1.5, 0.8, 3.1),
  pvalue = c(0.001, 0.04, 0.35, 0.0001),
  stringsAsFactors = FALSE   # keep characters as characters!
) )
```

\normalsize

:::: {.callout-warning}
Prior to R 4.0, `data.frame()` automatically converted character columns to **factors**. Always use `stringsAsFactors = FALSE` unless you specifically need factors.
::::

## Data Frame Subsetting

Data frames use `[row, column]` for subsetting:

\footnotesize

```{r}
# Select rows 1-2
dat[1:2, ]

# Select a column (returns a vector by default)
dat[, "gene"]

# Select by condition
dat[dat$pvalue < 0.05, ]     # significant genes only
```

## Data Frame — Adding & Removing Columns

\footnotesize

```{r}
# Add a column
dat$significant <- dat$pvalue < 0.05
dat

# Add with cbind()
cbind(dat, rank = 1:4)

# Remove a column
dat$significant <- NULL
dat
```

\normalsize

You can also append rows with `rbind()`, but column names must match.

## Tibble — The Modern Data Frame

A **tibble** is the tidyverse's upgraded data frame. Use `tibble()` or `tribble()`:

\footnotesize

```{r}
library(tibble)

# tibble() — by column
( tb <- tibble(
  gene  = c("BRCA1", "TP53", "EGFR"),
  logFC = c(2.3, -1.5, 0.8),
  sig   = c(TRUE, FALSE, TRUE)
) )
```

## Tibble — The Modern Data Frame, cont.

\footnotesize

```{r}
# tribble() — by row (easier to read for small data)
tribble(
  ~gene,  ~logFC, ~sig,
  "BRCA1", 2.3,  TRUE,
  "TP53", -1.5,  FALSE,
  "EGFR",  0.8,  TRUE
)
```


## Tibble vs Data Frame — Key Differences

### 1. Sequential column evaluation

\footnotesize

```{r}
# Tibble: columns evaluated left-to-right
tibble(x = 1:5, y = x^2)

# data.frame: this fails!
# data.frame(x = 1:5, y = x^2)
```

## Tibble vs Data Frame — Key Differences, cont.

### 2. Subsetting never simplifies

\footnotesize

```{r}
df <- data.frame(x = 1:3, y = 3:1)
tb <- tibble(x = 1:3, y = 3:1)

# data.frame: single column becomes a vector
class(df[, 1])              # "integer"

# tibble: always stays a tibble
class(tb[, 1])              # "tbl_df" "tbl" "data.frame"
```

## Tibble vs Data Frame — More Differences

### 3. Recycling rules

\footnotesize

```{r}
# data.frame: recycles freely (6 is multiple of 2)
data.frame(a = 1:6, b = LETTERS[1:2])     # works silently!
```

## Tibble vs Data Frame — More Differences, cont.

\footnotesize

```{r}
# tibble: only length-1 or exact match
# tibble(a = 1:6, b = LETTERS[1:2])       # ERROR!
tibble(a = 1:6, b = 1)                    # length-1: OK
```

### 4. Partial matching

\footnotesize

```{r}
df <- data.frame(abc = 1:3)
df$ab                       # partial match — dangerous!

tb <- tibble(abc = 1:3)
# tb$a                       # NULL + warning — safer!
```

## Converting Between Types

\footnotesize

```{r}
# data.frame → tibble
as_tibble(iris)

```

## Converting Between Types, cont.

\footnotesize
```{r}
# tibble → data.frame
as.data.frame(as_tibble(iris)[1:5, ])
```

\normalsize

:::: {.callout-tip}
In this course, we prefer **tibbles** for everyday work. Convert to `data.frame` only when a function requires it.
::::

## Working with tibbles — Add Rows & Columns

\footnotesize

```{r}
tb <- tibble(gene = character(), logFC = double())

# Add rows one by one
tb <- add_row(tb, gene = "BRCA1", logFC = 2.3)
tb <- add_row(tb, gene = "TP53",  logFC = -1.5)
tb

# Add a column
tb <- add_column(tb, significant = c(TRUE, FALSE))
tb


```

## Working with tibbles — Add Rows & Columns, cont.

\footnotesize

```{r}
# Combine tibbles by row
bind_rows(tb, tb)
```

## `attach()`, `with()`, and `within()`

\footnotesize

```{r}
# attach: access columns directly (use sparingly!)
attach(iris)
head(Sepal.Length, 3)
detach(iris)

# with: safer temporary access
with(iris, head(Sepal.Length, 3))

# within: modify multiple columns at once
head(within(airquality, {
  lOzone <- log(Ozone)
  cTemp  <- round((Temp - 32) * 5 / 9, 1)
}), 3)
```

# Data Import & Export

## Reading Data with `readr`

`readr` (part of tidyverse) provides fast, consistent file reading:

\footnotesize

| Function | Purpose |
|----------|---------|
| `read_csv()` | Comma-separated files |
| `read_tsv()` | Tab-separated files |
| `read_delim()` | Custom delimiters |
| `read_fwf()` | Fixed-width files |
| `read_table()` | Whitespace-separated columns |
| `read_log()` | Web server logs |

```{r eval=FALSE}
library(tidyverse)    # readr is included
# or: library(readr)
```

## Reading CSV Files

\footnotesize

```{r}
# Basic reading
myiris <- read_csv("data/talk03/iris.csv")
myiris
```

## Reading CSV Files, cont.

\footnotesize

With explicit column types:

```{r}
read_csv("data/talk03/iris.csv", col_types = cols(
  Sepal.Length = col_double(),
  Sepal.Width  = col_double(),
  Petal.Length = col_double(),
  Petal.Width  = col_double(),
  Species      = col_character()
))
```

## Column Type Specifications

\footnotesize

| `col_*()` | Type |
|-----------|------|
| `col_double()` | Double-precision numbers |
| `col_integer()` | Integers |
| `col_character()` | Text strings |
| `col_logical()` | TRUE/FALSE |
| `col_factor()` | Factors (categorical) |
| `col_date()` | Dates |
| `col_datetime()` | Date-times |
| `col_skip()` | Skip this column |
| `col_guess()` | Let readr guess |

\normalsize

Use `col_types = cols(...)` to ensure correct parsing, especially for date columns or numeric columns with special formatting.

## Writing Data

\footnotesize

| Function | Output format |
|----------|--------------|
| `write_csv()` | Comma-separated |
| `write_tsv()` | Tab-separated |
| `write_delim()` | Custom delimiter |
| `write_excel_csv()` | CSV optimised for Excel |
| `write_rds()` | R binary format |
| `write_lines()` | One string per line |

```{r eval=FALSE}
# Write iris to CSV
write_csv(iris, "iris_output.csv")

# Write to TSV
write_tsv(iris, "iris_output.tsv")
```

## Reading Other Formats

\footnotesize

| Package | Format |
|---------|--------|
| **readxl** | Excel (.xls, .xlsx) |
| **haven** | SPSS, Stata, SAS |
| **jsonlite** | JSON |
| **xml2** | XML |
| **DBI** | Databases (SQLite, PostgreSQL) |
| **httr** | Web APIs |
| **rvest** | HTML (web scraping) |

\normalsize

For bioinformatics: use `read_csv()`/`read_tsv()` for most flat files; `readxl` for Excel files from collaborators; `httr` for NCBI/Ensembl APIs.

# Tidy Data

## What is Tidy Data?

**Tidy data** is a standardized format where:

1.  Each **variable** has its own column
2.  Each **observation** has its own row
3.  Each **value** has its own cell

\vspace{0.3cm}

:::: {.callout-tip}
Tidy data makes it easy to filter, group, summarize, and plot. It's the foundation of the entire tidyverse workflow. (See R4DS by Wickham & Grolemund for visual examples.)
::::

## Tidy vs Untidy — Example

**Untidy** (wide format — common in Excel):

\footnotesize

```{r}
# Column headers are values (genes), not variable names
untidy <- tibble(
  patient = c("P1", "P2", "P3"),
  BRCA1   = c(2.3, 1.8, 3.1),
  TP53    = c(-1.2, -0.8, 0.5),
  EGFR    = c(0.5, 1.1, -0.3)
)
untidy
```

## Tidy vs Untidy — Example, cont.

**Tidy** (long format):

\footnotesize

```{r}
library(tidyr)
tidy <- pivot_longer(untidy, cols = c(BRCA1, TP53, EGFR),
                     names_to = "gene", values_to = "expression")
tidy
```

## Why Tidy Data?

1.  **dplyr** and **ggplot2** are designed for tidy data
2.  Easier to filter, group, and summarize
3.  Consistent mental model — you always know where to find things
4.  The foundation of the entire tidyverse workflow

\vspace{0.3cm}

:::: {.callout-tip}
> "Tidy datasets are all alike, but every messy dataset is messy in its own way."
> — Hadley Wickham
::::

# Dates with lubridate

## Working with Dates — `lubridate`

Biological data often involves dates: sample collection dates, patient birth dates, follow-up visits. The **lubridate** package makes date handling intuitive.

\footnotesize

```{r}
library(lubridate)

# Parse dates from strings
ymd("2024-03-15")            # year-month-day
mdy("March 15, 2024")        # month-day-year
dmy("15-03-2024")            # day-month-year

# Parse date-times
ymd_hms("2024-03-15 14:30:00")
```

## Date Extraction & Arithmetic

\footnotesize

```{r}
today()                       # current date
now()                         # current date-time

# Extract components
date <- ymd("2024-03-15")
year(date)                    # 2024
month(date)                   # 3
day(date)                     # 15
wday(date, label = TRUE)      # Fri (weekday name)
```

## Date Extraction & Arithmetic, cont.

\footnotesize

```{r}
# Date arithmetic
date + days(7)                # one week later
date - days(30)               # 30 days earlier
```

# Summary & Homework

## Today's Summary

\footnotesize

| Topic | Key Concepts |
|-------|-------------|
| **Factors** | `factor()`, `levels`, `ordered`, `forcats` (`fct_relevel`, `fct_recode`, `fct_lump`) |
| **Lists** | Flexible containers, `$` / `[[ ]]` vs `[ ]`, nested access |
| **Data Frames** | `data.frame()`, `nrow`/`ncol`, subsetting `[row, col]`, `rbind`/`cbind` |
| **Tibbles** | `tibble()`/`tribble()`, no simplification, strict recycling, sequential evaluation |
| **File I/O** | `read_csv`, `read_tsv`, `write_csv`, `col_types` |
| **Tidy Data** | Variables = columns, observations = rows, `pivot_longer` |
| **lubridate** | `ymd()`, `mdy()`, `today()`, date arithmetic |

## Homework

\footnotesize

1. Open `Homework_AI_powered_R/talk03-homework.qmd` in your course repository
2. Complete all exercises — use **CodeBuddy AI** for help!
3. Render to PDF
4. Submit as instructed on the course platform

:::: {.callout-tip}
**AI tips for this talk:**
- "Create a factor with levels in this specific order: ..."
- "How do I extract a nested element from a list in R?"
- "Why does my `read_csv` parse this column as character instead of numeric?"
- "Convert this wide gene expression table to tidy long format"
::::

## Next Talk

**Talk 04 — Data Wrangling with tidyverse**

- The pipe operator (`%>%` / `|>`)
- `dplyr` core verbs: `filter`, `select`, `mutate`, `arrange`, `summarise` + `group_by`
- `tidyr`: `pivot_longer`, `pivot_wider`, `separate`, `unite`
- Joins: `left_join`, `inner_join`, `full_join`
