---
title: "R for bioinformatics, data summarisation and statistics"
subtitle: "HUST Bioinformatics course series"
author: "Wei-Hua Chen (CC BY-NC 4.0)"
institute: "HUST, China"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  beamer_presentation:
    theme: AnnArbor
    colortheme: beaver
    fonttheme: structurebold
    highlight: tango
    includes:
      in_header: mystyle.sty
---

```{r include=FALSE}
color_block = function(color) {
  function(x, options) sprintf('\\color{%s}\\begin{verbatim}%s\\end{verbatim}',
                               color, x)
}

## 将错误信息用红色字体显示
knitr::knit_hooks$set(error = color_block('red'))
```

# section 1: TOC

## 前情提要

* basic plot functions
* basic ggplot2 
* special letters
* equations
* advanced ggplot2

## 本次提要

* data summarisation functions (vector data)
  * median, mean, sd, quantile, summary
* 图形化的 data summarisation (two-D data/ tibble/ table)
  * dot plot
   * smooth
   * linear regression
   * correlation & variance explained
   * groupping & bar/ box/ plots
* statistics
   * parametric tests
     * t-test
     * one way ANNOVA
     * two way ANNOVA
     * linear regression
     * model / prediction / coefficients
   * non-parametric comparison
   
# section 2: vector summarisation 


## test2 

\FontSmall

```{r}
wts2 <- bind_rows( 
   tibble( class = 1, age = sample( 13:15, 20, replace = T ), wt = sample( seq(50, 60, by = 0.1), 20 ) ),
   tibble( class = 2, age = sample( 14:16, 20, replace = T ), wt = sample( seq(55, 65, by = 0.1), 20 ) ),
   tibble( class = 3, age = sample( 15:17, 20, replace = T ), wt = sample( seq(60, 70, by = 0.1), 20 ) )
);

```


## test

\FontSmall

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(relaimpo);
```

```{r eval=FALSE}
library(relaimpo);
```

```{r echo=FALSE}
res3 <- calc.relimp( wt ~ factor(class) + age, data = wts2 );
res3$R2; ## 总 R2
res3$lmg; ## 每个因素的贡献； 

## 测试 rela 参数：
res4 <- calc.relimp( wt ~ factor(class) + age, data = wts2, rela = T);
res4$R2; ## 总 R2
res4$lmg; ## 每个因素的贡献； 

```