---
title: "AI-powered bioinformatic analysis using R, #7"
subtitle: "Data Visualization III — Bioinformatics Plots with Real Data"
author: "Wei-Hua Chen (CC BY-NC 4.0)"
institute: "HUST, China"
date: today

CJKmainfont: "Arial"

format: 
  beamer:
    aspectratio: 1610
    navigation: vertical
    theme: AnnArbor
    colortheme: beaver
    fonttheme: structurebold
    highlight: tango
    slide_level: 2
    toc: true
    pdf-engine: xelatex
    header-includes:
      - \AtBeginSubsection{}

execute:
  echo: true
  fig-width: 8
  fig-height: 4.5
---

# Bioinformatics Visualization Overview

## Common Plot Types in Bioinformatics

| Plot Type | Use Case | Typical Data |
|-----------|----------|--------------|
| **Stacked bar** | Microbiome composition | OTU/ASV abundance table |
| **Heatmap** | Gene expression, abundance patterns | Matrix of features × samples |
| **PCA / PCoA** | Sample clustering, batch effects | Any multivariate data |
| **Volcano plot** | Differential expression/abundance | Fold change + p-value |
| **Box/Violin plot** | Group comparisons | Alpha diversity, gene expression |
| **MA plot** | Expression vs. mean | RNA-seq count data |
| **Manhattan plot** | GWAS results | SNP positions + p-values |
| **Phylogenetic tree** | Evolutionary relationships | Sequence alignments |

## Today's Focus

We'll create **real bioinformatics figures** using actual microbiome and genomics data, with proper interpretation.

# Stacked Bar Plots for Microbiome

## Microbiome Data Structure

\footnotesize
```{r}
library(tidyverse)

# Mock microbiome species abundance data
speabu <- read_tsv(file = "data/talk09/mock_species_abundance.txt")
head(speabu)
```
\normalsize

Each row: one genus in one sample. Columns: sample `id`, `genus`, relative `abundance` (%).

## Stacked Bar Plot

\footnotesize
```{r fig.width=10, fig.height=4}
ggplot(speabu, aes(x = id, y = abundance, fill = genus)) +
  geom_bar(stat = "identity", position = "stack", color = "black", width = 0.8)
```
\normalsize

**Interpretation**: Each bar = one sample; each color segment = one genus. Compare community composition across samples.

## Control Genus Order with reorder()

\footnotesize
```{r fig.height=4, fig.width=10}
speabu$genus <- reorder(speabu$genus, speabu$abundance, median)
ggplot(speabu, aes(x = id, y = abundance, fill = genus)) +
  geom_bar(stat = "identity", position = "stack", color = "white", width = 0.8)
```
\normalsize

`reorder()` sorts genera by median abundance — high-abundance genera at top.

## position = "dodge"

\footnotesize
```{r fig.height=4, fig.width=10}
ggplot(speabu, aes(x = id, y = abundance, fill = genus)) +
  geom_bar(stat = "identity", position = "dodge", color = "white", width = 0.8)
```
\normalsize

Dodge = side-by-side bars. Better for comparing individual genera across samples.

## Add Labels on Stacked Bars

\footnotesize
```{r fig.height=4, fig.width=10}
speabu <- speabu %>% arrange(id, desc(factor(genus))) %>%
  group_by(id) %>% mutate(ypos = cumsum(abundance) - abundance / 2)

ggplot(speabu, aes(x = id, y = abundance, fill = genus)) +
  geom_bar(stat = "identity", position = "stack", color = "black", width = 0.8) +
  geom_text(aes(y = ypos, label = paste(abundance, "%", sep = "")), color = "white")
```
\normalsize

## geom_bar() Default Stat

**Key insight**: `geom_bar(stat = "count")` counts occurrences. For pre-computed values, use `stat = "identity"`.

Other default stats:
- `geom_bar`: count
- `geom_boxplot`: boxplot
- `geom_histogram`: bin
- `geom_density`: density
- `geom_quantile`: quantile

# Real Microbiome Data

## Load Real Genus-level Abundance Data

\footnotesize
```{r}
# Real gut microbiome genus data (Qin et al.)
qin_data <- read_csv("data/talk12/data/Qin.genus.data.csv")
dim(qin_data)
qin_data[1:5, 1:8]  # first 5 samples, first 8 genera
```
\normalsize

This dataset contains ~250 samples × ~140 genera from a human gut microbiome study.

## Reshape to Long Format for ggplot2

\footnotesize
```{r}
# Convert from wide (genera as columns) to long format
qin_long <- qin_data %>%
  pivot_longer(-1, names_to = "genus", values_to = "abundance") %>%
  rename(sample = 1)

head(qin_long)
```
\normalsize

## Top Genera Stacked Bar

\footnotesize
```{r fig.height=5, fig.width=10}
# Identify top 8 genera by mean abundance
top_genera <- qin_long %>%
  group_by(genus) %>%
  summarise(mean_ab = mean(abundance)) %>%
  slice_max(mean_ab, n = 8) %>%
  pull(genus)

# Combine rare genera into "Others"
qin_plot <- qin_long %>%
  mutate(genus = ifelse(genus %in% top_genera, genus, "Others")) %>%
  group_by(sample, genus) %>%
  summarise(abundance = sum(abundance), .groups = "drop")

ggplot(qin_plot, aes(x = sample, y = abundance, fill = genus)) +
  geom_bar(stat = "identity", position = "stack") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance (%)",
       title = "Gut Microbiome Composition (Top 8 Genera)")
```
\normalsize

## Interpretation

- **Bacteroides** dominates in most samples — typical of Western gut microbiomes
- **Prevotella** is abundant in a subset of samples — enterotype indicator
- **Faecalibacterium** is consistently present — beneficial butyrate producer
- Large inter-individual variation is normal in microbiome studies

# PCA for Sample Clustering

## Why PCA in Bioinformatics?

- **Dimensionality reduction**: 140 genera → 2 principal components
- **Visualize sample relationships**: similar microbiomes cluster together
- **Detect batch effects**: samples from same batch clustering separately
- **Identify outliers**: samples far from the main cluster

## Prepare Data for PCA

\footnotesize
```{r}
# Remove low-abundance genera (present in <10% of samples)
qin_mat <- qin_data %>%
  column_to_rownames(colnames(qin_data)[1]) %>%
  as.matrix()

# Filter: genera present in at least 10% of samples
qin_filtered <- qin_mat[, colMeans(qin_mat > 0) >= 0.1]

# Log-transform (add pseudocount to avoid log(0))
qin_log <- log10(qin_filtered + 0.01)

dim(qin_log)
```
\normalsize

## Compute PCA

\footnotesize
```{r}
pca_res <- prcomp(t(qin_log), scale. = TRUE)

# Variance explained
var_exp <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

# Create plotting data frame
pca_df <- data.frame(
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2],
  sample = rownames(pca_res$x)
)

head(pca_df)
```
\normalsize

## PCA Plot — ggplot2

\footnotesize
```{r fig.height=4, fig.width=6}
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 2, alpha = 0.7, color = "steelblue") +
  labs(x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"),
       title = "PCA of Gut Microbiome Samples") +
  theme_minimal()
```
\normalsize

## PCA Interpretation

- **PC1** explains the most variance (typically 15-30% in microbiome data)
- Samples close together = similar microbiome composition
- Look for clusters that might correspond to:
  - Enterotypes (Bacteroides vs. Prevotella dominated)
  - Disease status (if metadata available)
  - Technical batches

# Box Plots & Violin Plots

## Alpha Diversity Comparison

In microbiome studies, **alpha diversity** measures within-sample diversity. Common metrics:
- **Richness**: number of species observed
- **Shannon index**: richness + evenness
- **Simpson index**: probability two random reads are from different species

## Simulated Diversity Data

\footnotesize
```{r}
set.seed(123)
div_data <- tibble(
  group = rep(c("Healthy", "Disease"), each = 30),
  Shannon = c(rnorm(30, mean = 3.5, sd = 0.5),
              rnorm(30, mean = 2.8, sd = 0.6)),
  Richness = c(rnorm(30, mean = 120, sd = 25),
               rnorm(30, mean = 85, sd = 30))
)
head(div_data)
```
\normalsize

## Box Plot with Significance

\footnotesize
```{r fig.height=3.5, fig.width=5}
ggplot(div_data, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_manual(values = c("Healthy" = "#4DBBD5", "Disease" = "#E64B35")) +
  labs(x = "", y = "Shannon Diversity Index",
       title = "Alpha Diversity: Healthy vs. Disease") +
  theme_minimal() +
  theme(legend.position = "none")
```
\normalsize

## Violin + Box Plot Combo

\footnotesize
```{r fig.height=3.5, fig.width=5}
ggplot(div_data, aes(x = group, y = Shannon, fill = group)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("Healthy" = "#4DBBD5", "Disease" = "#E64B35")) +
  labs(x = "", y = "Shannon Diversity Index",
       title = "Distribution Comparison") +
  theme_minimal() +
  theme(legend.position = "none")
```
\normalsize

**Interpretation**: The disease group shows lower Shannon diversity (p < 0.001 by t-test). Reduced gut microbiome diversity is associated with many diseases.

# Heatmaps

## When to Use Heatmaps

Heatmaps visualize **matrix data** — rows = features (genes, genera), columns = samples, color = expression/abundance.

Common in:
- Gene expression clustering
- Microbiome abundance patterns
- Correlation matrices

## Simple Heatmap with ggplot2

\footnotesize
```{r fig.height=5, fig.width=10}
# Select top 20 most abundant genera
top20 <- qin_long %>%
  group_by(genus) %>%
  summarise(mean_ab = mean(abundance)) %>%
  slice_max(mean_ab, n = 20) %>%
  pull(genus)

# Prepare matrix subset
qin_heatmap <- qin_long %>%
  filter(genus %in% top20) %>%
  mutate(abundance = log10(abundance + 0.01))

ggplot(qin_heatmap, aes(x = sample, y = genus, fill = abundance)) +
  geom_tile() +
  scale_fill_gradient(low = "navy", high = "yellow",
                      name = "log10(Abundance)") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "Sample", y = "Genus",
       title = "Gut Microbiome Heatmap (Top 20 Genera)")
```
\normalsize

## Heatmap Interpretation

- **Rows**: individual genera (features)
- **Columns**: individual samples
- **Color intensity**: relative abundance (yellow = high, dark = low)
- **Patterns**:
  - Vertical stripes = samples with similar genus profiles
  - Horizontal bands = genera consistently high/low across samples
  - Clustered patterns suggest co-occurring genera

---

# Summary

## What We Covered

| Plot Type | Key Function | Bioinformatics Use |
|-----------|-------------|-------------------|
| **Stacked Bar** | `geom_bar(stat="identity")` | Microbiome composition |
| **PCA** | `prcomp()` + `geom_point()` | Sample clustering |
| **Box/Violin** | `geom_boxplot()`, `geom_violin()` | Group comparisons (diversity) |
| **Heatmap** | `geom_tile()` | Feature × sample matrices |
| **Position** | `"stack"`, `"dodge"`, `"jitter"` | Bar/point layout control |

## Key Takeaway

Bioinformatics visualization is about **telling the data's story**: composition, clustering, comparison, and patterns — all with clear interpretation.

## Next Talk

**Talk 08**: Strings, Iteration & Functional Programming (`stringr` + `purrr` + apply family)

## Homework

Complete `Homework_AI_powered_R/talk07-homework.qmd`
