---
title: "R for bioinformatics, data visualisation, part 2"
subtitle: "HUST Bioinformatics course series"
author: "Wei-Hua Chen (CC BY-NC 4.0)"
institute: "HUST, China"
date: today

CJKmainfont: "Hei"  # 关键：指定中文字体为黑体

format: 
  beamer:
    aspectratio: 1610 ## more morden aspectratio --
    navigation: vertical
    theme: AnnArbor
    colortheme: beaver
    fonttheme: structurebold
    highlight: tango
    slide_level: 2
    toc: true
  pdf:
    pdf-engine: xelatex  # 关键：指定XeLaTeX引擎

execute:
  echo: true
---

# section 1: TOC

## 前情提要

### iterations 与 并行计算

-   for loop
-   `apply` functions
-   `dplyr` 的本质是 遍历
-   `map` functions in `purrr` package
-   遍历 与 并行计算

### 相关包

-   purrr
-   parallel
-   foreach
-   iterators

## 前情提要, cont.

### ggplot2 basics

-   图层 ( geom_xxx )
-   scale ( scale_xxx )
-   坐标系统
-   faceting ( facet_xxx )

## 本次提要

### part 2： ggplot2 进阶

1.  如何在一张图中画多个panel？
2.  如何写公式
3.  ggplot2的核心是先计算，再做图
4.  themes and legends
5.  stacked bars 及其它

# ggplot2 进阶 1 ：如何在一张图中画多个panel？

## key requirements for multi-panel plots

-   order / position
-   labeling
-   layout

## combine multiple plots

Useful packages:

-   `gridExtra`
-   `cowplot`
-   `grid`
-   `lattice`

\footnotesize

install or load packages

```{r message=FALSE}
if (!require("gridExtra")){ 
  install.packages("gridExtra");
} 

if (!require("cowplot")){ 
  install.packages("cowplot");
} 

library( cowplot );
library( gridExtra );
```

## arranging multiple graphs using `cowplot`

### cowplot:

1.  `plot_grid`
2.  `ggdraw` + `draw_plot`

## prepare four panels

\footnotesize

```{r message=FALSE}
library(tidyverse);
plot.hwy <- ggplot(mpg, aes(x = cty, y = hwy, colour = factor(cyl)))+ 
  geom_point(size=2.5)
# Bar plot
plot.dose <- ggplot(diamonds, aes(clarity, fill = cut)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=70, vjust=0.5))
plot.iris <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point() + facet_grid(. ~ Species) + stat_smooth(method = "lm") +
  background_grid(major = 'y', minor = "none") + # add thin horizontal lines 
  panel_border();
plot.swiss <- swiss %>% ggplot(aes(Education, Agriculture)) +
  geom_point() + geom_smooth();
```

## Combine four plots using plot_grid

\footnotesize

```{r fig.height=5, fig.width=10, message=FALSE}
cowplot::plot_grid(plot.hwy, plot.dose, plot.iris, plot.swiss,
                   labels=c("A", "B", "C", "D"), ncol = 2, nrow = 2)
```

## 总结，本节内容

### ggplot2 基础

-   优缺点
-   用法
-   基本组成

### ggplot2 进阶

-   颜色和色板
-   复杂layout的实现
-   公式
-   ggplot2的数据统计逻辑

### 更多阅读

-   Ggplot2: Elegant Graphics for Data Analysis, Book by Hadley Wickham
-   ggplot2 gallery provided by RStudio on Github

## 写在后面

1.  ggplot2 博大精深，需要一门课去讲
2.  上手容易，精通难
3.  太多记忆点
4.  本节内容只涉及了基础中的基础，更多内容，包括进阶技巧和生信相关的扩展包，更多的需要同学们自行探索
5.  遇到不会的图，先百度/Google，找包和代码

## 下次预告

data summary and modeling

## 作业

-   `Exercises and homework` 目录下 `talk09-homework.Rmd` 文件；

-   完成时间：见钉群的要求