---
title: "Talk 08 — Homework"
subtitle: "Strings, Iteration & Functional Programming"
author: "Your Name — Your Student ID"
date: today
CJKmainfont: "Hei"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    pdf-engine: xelatex
  html:
    df-print: paged
---

## Instructions 说明

1.  Run each code chunk to verify results. 运行每个代码块以验证结果。
2.  For string exercises, use `stringr` functions. For iteration, use both base R and `purrr` where appropriate.
3.  Submit the compiled PDF as `姓名-学号-talk08-homework.pdf`.

## User Verification 用户验证

```{r}
Sys.info()[["user"]]
Sys.getenv("HOME")
```

```{r}
library(tidyverse)
library(stringr)
```

---

# Exercise 1: String Basics 字符串基础 {.unnumbered}

## 1.1 Basic Operations

Using `x <- c('weihua', 'chen')`:

1. Get the length of each element 每个元素的长度
2. Capitalize the first letter of each 首字母大写
3. Extract the first two characters of each 取前两个字符
4. Combine into one string separated by `', '` 合并为一个字符串
5. Count the number of **vowels** (a, e, i, o, u) in each element 数元音字母

```{r}
x <- c('weihua', 'chen')

## Your code here

```

## 1.2 Working with mtcars Row Names

1. Filter all Mercedes-Benz cars 筛选奔驰车
2. Filter all non-Mercedes-Benz cars 筛选非奔驰车
3. Split rownames into brand and model. E.g., `"Mazda RX4 Wag"` → `c("Mazda", "RX4 Wag")` 拆分品牌和车型

```{r}
## Your code here

```

## 1.3 str_c with Prefix and Suffix

Add prefix `"|-"` and suffix `"-|"` to `x <- c("abc", NA)` so the result is: `"|-abc-|" "|-NA-|"`

```{r}
x <- c("abc", NA)

## Your code here

```

---

# Exercise 2: Regular Expression Practice 正则表达式 {.unnumbered}

## 2.1 starwars Queries

1. Select all characters whose `skin_color` contains `"white"`. Show `name`, `homeworld`, `species`, `skin_color`.
2. Print all names containing `"ar"` (case-insensitive).

```{r}
## Your code here

```

## 2.2 Working with vec

Using `vec <- c("123", "abc", "wei555hua666")`:

1. Find strings containing digits 找出含数字的字符串
2. Locate the position of the first digit in each string 定位第一个数字
3. Locate all digit positions 定位所有数字
4. Extract the first digit group 提取第一组数字
5. Extract all digit groups 提取所有数字
6. Replace digits with `"666"` 替换数字为666

```{r}
vec <- c("123", "abc", "wei555hua666")

## Your code here

```

## 2.3 DNA Sequence Analysis 序列分析

Using `seq <- "ATCTCGGCGCGCATCGCGTACGCTACTAGC"`:

1. Get its reverse complement 反向互补 (A↔T, C↔G)
2. Calculate GC content as percentage GC含量
3. Split into codons (groups of 3 nucleotides) 拆分为密码子

```{r}
seq <- "ATCTCGGCGCGCATCGCGTACGCTACTAGC"

## Your code here

```

## 2.4 str_pad

**Question**: What does `stringr::str_pad` do? Provide an example. `str_pad` 的作用是什么？举例说明。

**Answer**:

```{r}
## Your example code here

```

## 2.5 Repeated Pattern Extraction

Extract substrings of length 2 that repeat 2+ times from these strings. E.g., `"1212"` contains `"12"` repeated twice; `"898989"` contains `"89"` repeated 3 times.

`c("banana", "coconut", "1232323", "database")`

```{r}
## Your code here

```

## 2.6 Number Matching Regex

Design a regex that matches all of these number formats:

```
123
123.45
0.124
-1.5
-0.2
+1.3
-11
-199.62
```

```{r}
## Your code here

```

---

# Exercise 3: for Loops 循环 {.unnumbered}

## 3.1 Basic Loop Practice

1. Write a loop to compute and print squares of numbers 1 to 7 打印1到7的平方
2. Print each column name of `iris` with its length, in format: `Sepal.Length (12)` 打印列名及长度
3. Write a `while` loop that draws `rnorm(1)` each time until the value is > 1 直到取到>1的数
4. Write a loop to find how many Fibonacci numbers are needed before the value exceeds 1 million. Fibonacci: `0, 1, 1, 2, 3, 5, 8, ...` 计算Fibonacci超过100万需要的循环数

```{r}
## Your code here

```

---

# Exercise 4: apply Family {.unnumbered}

## 4.1 Matrix Operations

Create a 100 × 100 numeric matrix. Compute:

1. Row means using `rowMeans()` and `apply()`
2. Column means using `colMeans()` and `apply()`
3. Row sums using `rowSums()` and `apply()`
4. Column sums using `colSums()` and `apply()`
5. Using a custom function, compute row mean, sum, and sd simultaneously

```{r}
## Your code here

```

## 4.2 tapply and dplyr

Using `mtcars`:

1. Compute mean `mpg` by `cyl` using `tapply()`
2. Compute mean `wt` by `cyl` using `tapply()`
3. Reproduce both using `dplyr` (group_by + summarise)

```{r}
## Your code here

```

## 4.3 lapply and sapply

1. Using `lapply` and `sapply`, compute the length of each member in: `list(a = 1:10, b = letters[1:5], c = LETTERS[1:8])`
2. Using `lapply` and `sapply`, compute the mean of each column in `mtcars`

```{r}
## Your code here

```

---

# Exercise 5: purrr 函数式编程 {.unnumbered}

## 5.1 map Basics

Using `df <- tibble(a = rnorm(10), b = rnorm(10), c = rnorm(10), d = rnorm(10))`, compute column means, sums, and medians with `map`.

```{r}
df <- tibble(a = rnorm(10), b = rnorm(10), c = rnorm(10), d = rnorm(10))

## Your code here

```

## 5.2 map Advanced

For each `cyl` group in `mtcars`, compute the Pearson correlation between `mpg` and `wt`, extracting both the correlation coefficient and p-value.

```{r}
## Your code here

```

## 5.3 keep and discard

1. Keep only factor columns from `iris`, print first 10 rows
2. Discard factor columns from `iris`, print first 10 rows

```{r}
## Your code here

```

## 5.4 reduce

Use `reduce()` to find the common elements among:

```{r}
c(1, 3, 5, 6, 10)
c(1, 2, 3, 7, 8, 10)
c(1, 2, 3, 4, 8, 9, 10)
```

Also: reproduce `reduce(full_join)` using `pivot_wider`.

```{r}
## Your code here

```

---

# Exercise 6: map2 and pmap {.unnumbered}

## 6.1 map2

Run the code below and reproduce the result using a `for` loop.

```{r}
mu <- list(5, 10, -3)
sigma <- list(1, 5, 10)
map2(mu, sigma, rnorm, n = 5)
```

```{r}
## Reproduce with for loop

```

## 6.2 pmap

Run the code below and reproduce the result using a `for` loop.

```{r}
params <- tribble(
  ~mean, ~sd, ~n,
     5,    1,  1,
    10,    5,  3,
    -3,   10,  5
)
params %>% pmap(rnorm)
```

```{r}
## Reproduce with for loop

```

---

# Exercise 7: Custom Functions 自定义函数 {.unnumbered}

## 7.1 Write a Function

Write a function `summarize_numeric(df)` that:
1. Takes a data frame as input
2. Checks each column is numeric (`stopifnot`)
3. Returns a tibble with columns: `variable`, `n`, `mean`, `sd`, `min`, `max`
4. Handles NA values

Test it on `mtcars`.

```{r}
## Your code here

```

## 7.2 Write a DNA Function

Write a function `gc_content(seq)` that:
1. Takes a DNA sequence string
2. Validates it only contains A, T, C, G (case-insensitive)
3. Returns the GC content as a percentage (rounded to 2 decimal places)

```{r}
## Your code here

```

---

# Exercise 8: Parallel Computing 并行计算 {.unnumbered}

## 8.1 Setup and Test

Run the following code and answer the questions.

```{r}
library(parallel)
library(foreach)
library(iterators)

# How many CPUs?
(cpus <- parallel::detectCores())

# Create test data
d <- data.frame(x = 1:10000, y = rnorm(10000))

# Make a cluster
cl <- makeCluster(cpus - 1)

# Parallel computation
res <- foreach(row = iter(d, by = "row")) %dopar% {
  return(row$x * row$y)
}

# Close cluster
stopCluster(cl)

summary(unlist(res))
```

**Question**: How many CPU cores does your system have? How many were used for this task? 你的系统有多少个CPU核心？本次任务用了几个？

**Answer**:

---

# Exercise 9: AI-Assisted Practice AI 辅助练习 {.unnumbered}

## 9.1 Regex Help

Ask your AI tool: *"Write a regex that extracts email addresses from text, and show me how to use it with stringr."*

**AI Response**:

```

```

## 9.2 purrr Help

Ask your AI tool: *"When should I use map_df() instead of map()? Give me a bioinformatics example."*

**AI Response**:

```

```

## 9.3 Debug

If any exercise above gave you errors, paste the error message here and ask your AI tool for help. Record the solution.

**Error**:

```

```

**AI Solution**:

```

```
