---
title: "Talk 04 — Homework"
subtitle: "Data Wrangling with tidyverse — Pipe, dplyr, tidyr & Joins"
author: "Your Name — Your Student ID"
date: today

CJKmainfont: "Hei"

format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    pdf-engine: xelatex
  html:
    df-print: paged
---

## Instructions 作业说明 {.unnumbered}

- Write your R code inside `{r}` code chunks, run them, and verify the output. 将R代码填入 `{r}` 代码块中，运行并验证输出结果。
- After completing all exercises, click the **Render** button in RStudio to generate a PDF. 完成后，点击 RStudio 中的 **Render** 按钮生成 PDF。
- **Rename the PDF** to: `Name-StudentID-talk04-homework.pdf` and submit to the course platform. **将PDF重命名**为：`姓名-学号-talk04-homework.pdf`，提交到课程平台。
- **Use CodeBuddy AI** whenever you get stuck — it's part of the learning process! 遇到困难时请使用 **CodeBuddy AI** ——这是学习过程的一部分！

## User Verification 用户身份验证 {.unnumbered}

Run the following to verify your identity 运行以下代码验证你的身份:

```{r}
Sys.info()[["user"]]
Sys.getenv("HOME")
getwd()
```

## Talk 04 Content Review 内容回顾

Today we covered 今天的内容包括:

- **The Pipe** 管道操作符: `%>%`, `.` placeholder, `%T>%`, `%$%`, `%<>%`
- **dplyr Core Verbs** 核心函数:
  - `select()` — choose columns 选择列
  - `filter()` — subset rows 过滤行
  - `mutate()` — create/modify columns 创建/修改列
  - `summarise()` + `group_by()` — grouped aggregation 分组汇总
  - `arrange()` — sort rows 排序
  - `across()` — column-wise operations 列操作
  - `rowwise()` — row-wise operations 行操作
- **tidyr**: `pivot_longer()` / `pivot_wider()`, `separate()` / `unite()`
- **Joins** 连接: `left_join`, `inner_join`, `full_join`, `semi_join`, `anti_join`

---

# Exercise 1: The Pipe 管道操作 {.unnumbered}

## 1.1 Rewrite nested code with pipes

Rewrite the following nested code using `%>%`. Verify both produce the same result. 用管道重写以下嵌套代码，验证结果相同。

```{r}
library(tidyverse)
# Original nested code
result_original <- head(arrange(summarise(group_by(filter(starwars, species == "Human"), gender), avg_height = mean(height, na.rm = TRUE)), desc(avg_height)), 3)
result_original
```

```{r}
# Your pipe version here
```

**Question**: Which version is easier to read and debug? Why? 哪个版本更易读易调试？为什么？

**Answer**:

## 1.2 Use `.` placeholder

Use the `.` placeholder to pass `mtcars` as a **non-first** argument:

```{r}
# Use %>% and . to pass mtcars to lm() as the data argument
# Fit: mpg ~ wt + hp
```

## 1.3 The tee pipe `%T>%`

Explain what happens in this code. Which value does `res` receive? 解释以下代码，`res` 最终接收到什么值？

```{r}
library(magrittr)
res <- rnorm(50) %>% matrix(ncol = 5) %T>% plot() %>% colMeans()
res
```

**Answer**:

---

# Exercise 2: dplyr Core Verbs 核心函数 {.unnumbered}

## 2.1 `select()` — choose columns

Using the `starwars` dataset 使用 `starwars` 数据集:

```{r}
library(tidyverse)

# (a) Select name, species, and all columns ending with "color"
# (b) Select columns from "name" through "mass" using column range
# (c) Remove the "films", "vehicles", and "starships" columns using -
```

## 2.2 `filter()` — subset rows

Using `starwars` 使用 `starwars`:

```{r}
# (a) Filter for Human characters with height > 180
# (b) Filter for characters whose homeworld is "Tatooine" OR "Naboo"
# (c) Filter for characters with BOTH blue eyes AND blond hair
# (d) Filter for characters with height between 150 and 200 (inclusive)
```

## 2.3 `mutate()` — create/modify columns

Using `starwars` 使用 `starwars`:

```{r}
# (a) Create a "bmi" column: mass / (height/100)^2
# (b) Create a "height_cat" column using case_when():
#     < 100 → "Short", 100–180 → "Medium", > 180 → "Tall"
# (c) Show only name, species, height, height_cat, bmi — sorted by bmi descending
```

## 2.4 `summarise()` + `group_by()` — grouped aggregation

Using `starwars` 使用 `starwars`:

```{r}
# (a) Count how many characters per species (use n()), sort by count descending
# (b) Calculate avg height and avg mass per gender (remove NAs)
# (c) Find the tallest character per species (use slice_max), only for species with ≥ 2 members
```

## 2.5 `arrange()` — sort rows

Using `starwars` 使用 `starwars`:

```{r}
# (a) Sort by mass descending, show the top 5 heaviest characters
# (b) Sort by species alphabetically, then by height descending within each species
```

---

# Exercise 3: Mouse Genes Analysis 小鼠基因分析 {.unnumbered}

## 3.1 Load and explore the data

Read the mouse genes file from `../data/talk04/mouse_genes_biomart_sep2018.txt` and explore: 读取小鼠基因数据并探索。

```{r}
# Load with read_delim() (tab-separated, quote = "")
```

```{r}
# Count genes per chromosome/scaffold, sorted by count descending
```

```{r}
# Count genes per transcript type, sorted by count descending
```

## 3.2 Filter and summarize

Complete the following analysis pipeline 完成以下分析管道:

```{r}
# 1. Keep only chromosomes 1–19, X, Y
# 2. Keep only "protein_coding", "miRNA", "lincRNA" transcript types
# 3. Rename columns for readability (short names)
# 4. Group by chromosome and type
# 5. Summarize: count of distinct genes, mean gene length, min and max gene length
# 6. Sort by chromosome (ascending), then count (descending)
```

```{r}
# Bonus: Remove chromosomes with fewer than 500 genes, 
# then re-sort by chromosome and count
```

## 3.3 Find extreme genes

For each chromosome, find the **longest** and **shortest** gene (by transcript length). Show gene ID, chromosome, and length. 找出每条染色体上最长和最短的基因。

```{r}
# Hint: use slice_max() and slice_min() after group_by()
```

---

# Exercise 4: Grades Analysis 成绩分析 {.unnumbered}

## 4.1 Create and reshape the data

Create the grades tibble and reshape to long format 创建成绩数据并转为长格式:

```{r}
set.seed(42)  # for reproducibility
grades <- tibble(
  "Name" = c("Weihua Chen", "Mm Hu", "John Doe", "Jane Doe",
             "Warren Buffet", "Elon Musk", "Jack Ma"),
  "Occupation" = c("Teacher", "Student", "Teacher", "Student",
                   rep("Entrepreneur", 3)),
  "English" = sample(60:100, 7),
  "ComputerScience" = sample(80:90, 7),
  "Biology" = sample(50:100, 7),
  "Bioinformatics" = sample(40:90, 7)
)

# Use pivot_longer() to reshape to long format (columns: Name, Occupation, course, grade)
```

## 4.2 Average grades

Calculate the average grade per student, sorted from highest to lowest. 计算每人平均成绩，从高到低排序。

```{r}
# Use group_by() + summarise() + arrange()
```

## 4.3 Best and worst subjects

Find each student's best subject and worst subject. 找出每人最强和最弱科目。

```{r}
# Best subject per student (use slice_max)
```

```{r}
# Worst subject per student (use slice_min)
```

## 4.4 By occupation

```{r}
# (a) Which occupation has the highest average grade?
# (b) For each occupation, which subject has the highest average score?
```

---

# Exercise 5: tidyr — Reshaping 数据重塑 {.unnumbered}

## 5.1 Wide to long

Convert this wide gene expression table to long (tidy) format 将宽格式转为长格式:

```{r}
expr_wide <- tibble(
  patient = c("P1", "P2", "P3", "P4"),
  treatment = c("Drug", "Drug", "Placebo", "Placebo"),
  IL6 = c(12.3, 15.1, 8.7, 9.2),
  TNF = c(5.6, 7.8, 3.2, 4.1),
  CRP = c(2.1, 3.4, 1.5, 1.8)
)

# Use pivot_longer() — keep patient and treatment as identifiers
```

## 5.2 Long to wide

Convert back to wide format 转回宽格式:

```{r}
# Use pivot_wider() on the long result from 5.1
```

## 5.3 Separate and unite

```{r}
# Data with combined drug-dose column
trial <- tibble(
  treatment = c("Aspirin_100mg", "Aspirin_200mg", "Ibuprofen_200mg", "Ibuprofen_400mg"),
  response = c(0.72, 0.85, 0.63, 0.78)
)

# (a) Use separate() to split treatment into drug and dose columns
```

```{r}
# (b) Use unite() to combine drug and dose back, separated by " - "
```

---

# Exercise 6: Joining Tables 表连接 {.unnumbered}

## 6.1 Basic joins

Two datasets from a clinical study 临床研究中的两个数据集:

```{r}
# Gene expression data
expr <- tibble(
  gene = c("BRCA1", "TP53", "EGFR", "MYC", "KRAS", "PTEN"),
  log2FC = c(2.3, -1.5, 0.8, 3.1, -0.4, -2.1),
  pvalue = c(0.001, 0.02, 0.15, 0.003, 0.08, 0.01)
)

# Gene annotation
annotation <- tibble(
  gene = c("BRCA1", "TP53", "EGFR", "MYC", "VEGFA", "PTEN"),
  chromosome = c("17", "17", "7", "8", "6", "10"),
  pathway = c("DNA repair", "Cell cycle", "Signaling", "Transcription", "Angiogenesis", "Signaling")
)

# (a) inner_join — keep only genes present in BOTH tables
```

```{r}
# (b) left_join — keep all genes from expr, add annotation where available
```

```{r}
# (c) full_join — keep ALL genes from both tables
```

## 6.2 Anti join — quality control

```{r}
# (a) Which genes are in expr but NOT in annotation?
# (b) Which genes are in annotation but NOT in expr?
```

## 6.3 Join interpretation

**Question**: In a bioinformatics analysis, you have a DEG table (differentially expressed genes) and a pathway table. You want to keep ALL DEGs and annotate them with pathways when available. Which join should you use? Why? 在一个生物信息学分析中，你有 DEG 表和 pathway 表。你想保留所有 DEG 并标注可用的 pathway，应该用哪种 join？为什么？

**Answer**:

---

# Exercise 7: Advanced dplyr 高级 dplyr {.unnumbered}

## 7.1 `across()` — column-wise operations

Using `starwars` 使用 `starwars`:

```{r}
# (a) Calculate the mean of ALL numeric columns (remove NAs)
# (b) For ALL numeric columns, create a centered version (value - mean, as new columns with "_centered" suffix)
```

## 7.2 `rowwise()` — row-wise operations

```{r}
# Gene expression: calculate the mean expression across samples per gene
gene_expr <- tibble(
  gene = c("BRCA1", "TP53", "EGFR", "MYC", "KRAS"),
  tumor1 = c(12.3, 8.7, 15.2, 22.1, 5.4),
  tumor2 = c(10.1, 9.3, 14.8, 20.5, 6.1),
  tumor3 = c(11.5, 7.9, 16.1, 21.3, 4.8),
  normal1 = c(5.2, 10.1, 14.0, 6.3, 5.0),
  normal2 = c(4.8, 9.5, 13.5, 5.8, 4.5)
)

# Use rowwise() + c_across() to calculate:
# - mean_tumor: mean of columns starting with "tumor"
# - mean_normal: mean of columns starting with "normal"
# - log2FC: log2(mean_tumor / mean_normal)
```

## 7.3 `slice_*()` — row selection

Using `Theoph` (built-in dataset) 使用内置 `Theoph` 数据集:

```{r}
data(Theoph)
# Convert to tibble
th <- as_tibble(Theoph)

# (a) Show the first 3 rows per Subject
# (b) Show the row with the maximum conc per Subject
# (c) Show the row with the minimum Time per Subject
```

---

# Exercise 8: Putting It All Together 综合练习 {.unnumbered}

## 8.1 Complete analysis pipeline

Using the `starwars` dataset, create a single pipeline that: 使用 `starwars` 完成以下综合管道:

```{r}
# 1. Filter to keep only characters with height and mass both > 0
# 2. Create a "bmi" column
# 3. Create a "bmi_cat" column: < 18.5 → "Underweight", 18.5–25 → "Normal", 
#    25–30 → "Overweight", > 30 → "Obese"
# 4. Group by species and bmi_cat
# 5. Count how many characters per species-bmi_cat combination
# 6. Keep only combinations with count ≥ 2
# 7. Arrange by species, then by bmi_cat (in order: Underweight → Normal → Overweight → Obese)
```

## 8.2 Join + reshape + summarize

```{r}
# Clinical trial data
patients <- tibble(
  patient_id = c("P01", "P02", "P03", "P04", "P05"),
  age = c(45, 62, 38, 55, 41),
  sex = c("F", "M", "F", "M", "F"),
  treatment = c("Drug", "Placebo", "Drug", "Drug", "Placebo")
)

# Gene expression (wide format)
expression <- tibble(
  patient_id = c("P01", "P02", "P03", "P04", "P05"),
  BRCA1 = c(2.3, 1.1, 3.5, 2.8, 1.9),
  TP53  = c(-1.2, 0.3, -2.1, -0.8, 0.1),
  EGFR  = c(0.5, -0.2, 1.8, 0.9, -0.5)
)

# (a) Join patients and expression by patient_id
# (b) Pivot expression to long format (gene, expression columns)
# (c) Group by treatment and gene, calculate mean expression
# (d) Pivot back to wide format: one row per gene, columns = treatment means
```

---

# Exercise 9: Use CodeBuddy to Help 使用 AI 辅助 {.unnumbered}

## 9.1 Debug with AI

If you encountered any errors in the exercises above, paste the error message to CodeBuddy and ask for help. Record the prompt and summary below. 如遇到错误，用 CodeBuddy 调试。

**My debug prompt**:

```
```

**CodeBuddy's response summary**:

## 9.2 Explain a concept with AI

Ask CodeBuddy: "Explain the difference between inner_join, left_join, and full_join in R dplyr. When should I use each in a bioinformatics context?"

**My prompt**:

```
```

**CodeBuddy's response summary**:

## 9.3 Generate extra practice

Ask CodeBuddy: "Give me 3 additional exercises about dplyr group_by and summarise with the starwars dataset, with answers."

**My prompt**:

```
```

**CodeBuddy's response summary**:

---

*Homework ends here. Make sure all code chunks run without errors, then Render to PDF and submit. 作业结束。请确认所有代码块运行无误，Render 生成 PDF 后提交。*
