---
title: "Talk 03 — Homework"
subtitle: "R Language Basics II — Factors, Lists, Data Frames, File I/O"
author: "Your Name — Your Student ID"
date: today

CJKmainfont: "Hei"

format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    pdf-engine: xelatex
  html:
    df-print: paged
---

## Instructions 作业说明 {.unnumbered}

- Write your R code inside `{r}` code chunks, run them, and verify the output. 将R代码填入 `{r}` 代码块中，运行并验证输出结果。
- After completing all exercises, click the **Render** button in RStudio to generate a PDF. 完成后，点击 RStudio 中的 **Render** 按钮生成 PDF。
- **Rename the PDF** to: `Name-StudentID-talk03-homework.pdf` and submit to the course platform. **将PDF重命名**为：`姓名-学号-talk03-homework.pdf`，提交到课程平台。
- **Use CodeBuddy AI** whenever you get stuck — it's part of the learning process! 遇到困难时请使用 **CodeBuddy AI** ——这是学习过程的一部分！

## User Verification 用户身份验证 {.unnumbered}

Run the following to verify your identity 运行以下代码验证你的身份:

```{r}
Sys.info()[["user"]]
Sys.getenv("HOME")
```

## Talk 03 Content Review 内容回顾

Today we covered 今天的内容包括:

- **Factors** 因子: `factor()`, `levels`, `ordered`, `forcats` (`fct_relevel`, `fct_recode`, `fct_lump`)
- **Lists** 列表: creation, naming, `$` / `[[ ]]` vs `[ ]`, nested access
- **Data Frames** 数据框: `data.frame()`, subsetting `[row, col]`, `rbind`/`cbind`, column operations
- **Tibbles** 数据表: `tibble()`/`tribble()`, vs data.frame differences, `add_row`/`add_column`/`bind_rows`
- **File I/O** 文件读写: `read_csv`, `read_tsv`, `write_csv`, `col_types`, `readr`
- **Tidy Data** 整洁数据: variables = columns, observations = rows, `pivot_longer`
- **lubridate** 日期处理: `ymd()`, `mdy()`, `today()`, date arithmetic

---

# Exercise 1: Factors 因子 {.unnumbered}

## 1.1 Create a basic factor

Create a factor from the following vector of blood types. Display the factor and its levels. 从以下血型向量创建因子，显示因子及其水平。

```{r}
blood_types <- c("A", "O", "B", "AB", "O", "A", "B", "A", "AB", "O")

## Write your code here — use factor()
```

## 1.2 Control the level order

Create the same factor but with levels in this order: `"O"`, `"A"`, `"B"`, `"AB"`. Why might you want to control level order? 用指定顺序创建因子：`"O"`, `"A"`, `"B"`, `"AB"`。为什么要控制水平顺序？

```{r}
## Write your code here
```

**Question**: Why would you choose this particular order (`O`, `A`, `B`, `AB`)? 为什么选择这个顺序？

**Answer**:

## 1.3 Ordered factor for disease severity

Create an **ordered** factor for disease severity: `"Mild"`, `"Severe"`, `"Moderate"`, `"Mild"`, `"Severe"`. Verify that `"Mild" < "Severe"` returns `TRUE`. 创建有序因子，验证比较运算。

```{r}
severity <- c("Mild", "Severe", "Moderate", "Mild", "Severe")

## Write your code here — use ordered() or factor(..., ordered = TRUE)
```

```{r}
## Test the comparison here
```

## 1.4 Count factor levels with `table()`

Use `table()` to count how many blood samples of each type are in the `blood_types` vector from 1.1. Which type is most common? 用 `table()` 统计每种血型的样本数。哪种血型最常见？

```{r}
## Write your code here
```

**Answer**:

## 1.5 `forcats` — reorder by frequency

Use `fct_infreq()` to reorder the levels of `blood_types` factor by frequency (most common first). Then use `table()` again to confirm. 用 `fct_infreq()` 按频率重新排序。

```{r}
library(forcats)

## Write your code here
```

## 1.6 `forcats` — rename factor levels

Use `fct_recode()` to rename the levels of the blood type factor: `"A"` → `"Type A"`, `"B"` → `"Type B"`, `"AB"` → `"Type AB"`, `"O"` → `"Type O"`. 用 `fct_recode()` 重命名因子水平。

```{r}
## Write your code here
```

## 1.7 Gene expression categories — practical exercise

Create an ordered factor from the gene expression data below. Levels should be: `"Down"`, `"Unchanged"`, `"Up"` (in that order). Then create a `table()` summary. 创建基因表达类别因子并做汇总。

```{r}
expr_result <- c("Up", "Up", "Down", "Unchanged", "Up", "Down",
                 "Unchanged", "Unchanged", "Up", "Down", "Down", "Up")

## Write your code here — create the factor with ordered levels
```

```{r}
## Write table() summary here
```

**Question**: How many genes are up-regulated? Down-regulated? Unchanged? 有多少基因上调、下调、不变？

**Answer**:

---

# Exercise 2: Lists 列表 {.unnumbered}

## 2.1 Create a patient record list

Create a list `patient` with these elements:
- `id`: `"P042"` (character)
- `age`: `58` (numeric)
- `diagnosis`: `"Type 2 Diabetes"` (character)
- `glucose_readings`: `c(142, 155, 138, 167, 149)` (numeric vector)
- `on_medication`: `TRUE` (logical)

创建一个包含病人信息的列表。

```{r}
## Write your code here
```

## 2.2 Access list elements

Using the `patient` list from 2.1, extract the following using **three different access methods** (`$`, `[[ ]]`, `[ ]`):

用三种不同的方式从列表中提取元素。

```{r}
# (a) Extract the patient id using $
# (b) Extract glucose_readings using [[ ]]
# (c) Extract the first two glucose readings using [ ] after [[ ]]
# (d) Use [ ] to get a sub-list containing only id and diagnosis
```

**Question**: What is the difference in result type between `patient["id"]` and `patient[["id"]]`? Use `class()` to verify. 两者结果类型有何不同？用 `class()` 验证。

**Answer**:

## 2.3 Nested list — differential expression result

Create a list `de_result` representing a differential expression analysis output:
- `gene`: `"TP53"`
- `statistics`: a **sub-list** with `log2FC = -2.1`, `p_value = 0.003`, `fdr = 0.042`
- `sample_ids`: `c("T1", "T2", "T3", "N1", "N2", "N3")`

Then extract the `fdr` value from the nested list using `$`. 创建嵌套列表并提取元素。

```{r}
## Write your code here
```

```{r}
## Extract fdr using $ chaining
```

## 2.4 Modify a list

Starting with the `patient` list from 2.1:
- Add a new element `bmi` with value `28.5`
- Change `diagnosis` to `"Type 2 Diabetes with complications"`
- Remove the `on_medication` element by setting it to `NULL`
- Display the modified list

从 2.1 的列表出发，添加、修改、删除元素。

```{r}
## Write your code here — modify patient list
```

## 2.5 List of analysis results

Create a list `analysis_results` that contains:
- Element `"pca"`: a numeric vector `c(PC1 = 0.45, PC2 = 0.23, PC3 = 0.12)`
- Element `"significant_genes"`: a character vector `c("BRCA1", "TP53", "EGFR", "MYC")`
- Use `length()` to check how many top-level elements the list has
- Use `length()` on `analysis_results$significant_genes` to see how many significant genes

创建分析结果列表。

```{r}
## Write your code here
```

---

# Exercise 3: Data Frames 数据框 {.unnumbered}

## 3.1 Create a data frame from scratch

Create a data frame `patients` with 4 patients:

| Name | Age | BloodType | Glucose |
|------|-----|-----------|---------|
| Zhang | 45 | A | 98 |
| Li | 62 | O | 142 |
| Wang | 38 | B | 105 |
| Chen | 55 | AB | 128 |

Use `stringsAsFactors = FALSE` to keep `BloodType` as character. 用 `data.frame()` 创建数据框，注意设置 `stringsAsFactors = FALSE`。

```{r}
## Write your code here
```

## 3.2 Inspect the data frame

Apply the following inspection functions to `patients` from 3.1 and report what you find: 用以下函数检查数据框。

```{r}
# Use: nrow(), ncol(), dim(), str(), summary()
```

## 3.3 Subset rows and columns

Using `patients` from 3.1, perform the following operations: 对数据框进行子集提取。

```{r}
# (a) Select the first two rows
# (b) Select the "Name" and "Glucose" columns
# (c) Select rows where Glucose > 110
# (d) Select the "BloodType" column as a vector (not a data frame)
```

## 3.4 Add and remove columns

Starting with `patients` from 3.1: 添加和删除列。

```{r}
# (a) Add a column "BMI" with values c(23.5, 28.1, 21.7, 25.3)
# (b) Add a column "Diabetic" as a logical: Glucose > 125
# (c) Remove the "BloodType" column
# (d) Display the result
```

## 3.5 Work with `VADeaths`

The built-in dataset `VADeaths` contains death rates in Virginia (1940). 使用内置数据集 `VADeaths`。

```{r}
# (a) Check the class of VADeaths
# (b) Convert it to a data frame
# (c) Add a "Total" column (sum of each row)
# (d) Reorder columns so "Total" is first
```

## 3.6 Work with `swiss`

The built-in `swiss` dataset contains Swiss fertility data. 使用内置 `swiss` 数据集。

```{r}
# (a) Select rows 1, 2, 3, 10, 11, 12, 13 and columns Examination, Education, Infant.Mortality
# (b) In the result, find the row named "Sarine" and set its Infant.Mortality to NA
# (c) Add a column "Mean" with the row-wise mean of each row
```

## 3.7 Merge three vectors into a data frame

Merge the following three vectors into a single data frame: 合并三个向量为数据框。

```{r}
Id <- LETTERS[1:26]
x  <- seq(1, 52, along.with = Id)
y  <- seq(-20, 20, along.with = Id)

## Write your code here
```

**Question**: What does `along.with` do in `seq()`? Explain with an example. `along.with` 参数的作用是什么？举例说明。

**Answer**:

## 3.8 Merge two data frames by a common column

Merge `df1` and `df2` by their `Id` column: 按共同列合并两个数据框。

```{r}
df1 <- data.frame(Id = 1:4, Age = c(14, 12, 15, 10))
df2 <- data.frame(Id = 1:4, Sex = c("F", "M", "M", "F"), Code = c("a", "b", "c", "d"))

## Write your code here — use merge() or cbind()
```

```{r}
# Then remove the "Code" column from the result
```

## 3.9 Recycling with `cbind()`

Run the following code and answer each question: 运行以下代码并回答问题。

```{r}
df3 <- data.frame(data = 1:10, group = c("A", "B"))

# (a) cbind(df3, newcol = 1)        — does it work?
# (b) cbind(df3, newcol = 1:10)     — does it work?
# (c) cbind(df3, newcol = 1:2)      — does it work?
# (d) cbind(df3, newcol = 1:3)      — does it work?
```

**Question**: Explain which cases work and which don't. What recycling rules apply to `cbind` with data frames? 哪些情况可行，哪些不可行？`cbind` 的循环规则是什么？

**Answer**:

---

# Exercise 4: Tibbles 数据表 {.unnumbered}

## 4.1 Create a tibble

Create a tibble with:
- `gene`: `c("BRCA1", "TP53", "EGFR", "MYC", "KRAS")`
- `log2FC`: `c(2.3, -1.5, 0.8, 3.1, -0.4)`
- `significant`: calculated as `abs(log2FC) > 1` (use the column reference!)

用 `tibble()` 创建数据表，利用 tibble 的顺序计算特性。

```{r}
library(tibble)

## Write your code here — create tibble with sequential column evaluation
```

## 4.2 Create a tibble with `tribble()`

Recreate the same gene data using `tribble()` (row-by-row format). 用 `tribble()` 按行创建同样的数据。

```{r}
## Write your code here
```

## 4.3 Tibble subsetting — test `drop` behavior

Create a tibble `tb <- tibble(x = 1:3, y = 3:1, z = c("a", "b", "c"))` and extract: 测试 tibble 子集行为。

```{r}
tb <- tibble(x = 1:3, y = 3:1, z = c("a", "b", "c"))

# (a) Extract column x — use class() to check if it stays a tibble
# (b) Extract column x as a vector (not a tibble)
# (c) Extract columns x and y as a tibble
```

**Question**: What is the key difference between `tb[1]`, `tb[[1]]`, and `tb$x`? Which returns a tibble vs a vector? 三种访问方式分别返回什么类型？

**Answer**:

## 4.4 Explain the recycling difference

Explain why this works: 解释为什么 data.frame 可行而 tibble 不可行。

```{r eval=FALSE}
data.frame(a = 1:6, b = LETTERS[1:2])   # works
# tibble(a = 1:6, b = LETTERS[1:2])     # doesn't work — why?
```

**Question**: What is the recycling rule difference between `data.frame()` and `tibble()`? data.frame 和 tibble 的循环规则有何不同？

**Answer**:

## 4.5 Build a tibble row-by-row

Create an empty tibble with columns `name` (character), `age` (integer), `score` (double), then use `add_row()` to add three rows of random data. 创建空 tibble 并逐行添加数据。

```{r}
## Write your code here
## Step 1: Create empty tibble with column definitions
## Step 2: Add three rows using add_row()
```

## 4.6 Convert between data frame and tibble

```{r}
# (a) Convert the built-in "mtcars" dataset to a tibble
# (b) Print the first 3 rows
# (c) Convert it back to a regular data frame
```

## 4.7 Tibble advantages summary

**Question**: List at least 4 advantages of `tibble` over `data.frame`. Show a code example for one of them. 列出 tibble 相比 data.frame 的至少4个优势，并用代码示例说明其中一个。

**Answer**:

```{r}
## Code example for one advantage
```

---

# Exercise 5: Data Import / Export 文件读写 {.unnumbered}

## 5.1 Read iris.csv with explicit column types

Read `../data/talk03/iris.csv` using `read_csv()`. Use `col_types` to specify all 5 columns explicitly (4 × `col_double()` + 1 × `col_character()`). 用 `read_csv()` 读取 iris.csv，显式指定列类型。

```{r}
library(tidyverse)

## Write your code here
```

```{r}
# Check the result with str() or glimpse()
```

## 5.2 Read a TSV file

Read `../data/talk03/states1.csv` (it's actually a CSV with semicolons as delimiter — check its content first). Note: some files use `;` or other delimiters. Show how to read it correctly with `read_delim()`. 读取使用非标准分隔符的文件。

```{r}
# First, use read_lines() to peek at the first 3 lines of the file
```

```{r}
# Then read it with the correct delimiter
```

## 5.3 Write data to file

Using the `patients` data frame from Exercise 3.1: 将数据框写入文件。

```{r}
# (a) Write to "patients.csv" using write_csv()
# (b) Write to "patients.tsv" using write_tsv()
# (c) Verify by reading the files back
```

## 5.4 Convert wide gene expression to long format

**Tidy data practice**: Convert the following wide gene expression table into tidy long format using `pivot_longer()`. 将宽格式基因表达表转换为整洁的长格式。

```{r}
library(tidyr)

wide_expr <- tibble(
  patient = c("P1", "P2", "P3"),
  BRCA1   = c(2.3, 1.8, 3.1),
  TP53    = c(-1.2, -0.8, 0.5),
  EGFR    = c(0.5, 1.1, -0.3)
)

## Write pivot_longer() here
```

**Question**: Why is the long format considered "tidy" for this data? What operations become easier? 为什么长格式被认为是"整洁的"？哪些操作变得更简单了？

**Answer**:

## 5.5 Read real-world files — overview 真实文件读取概述 {.unnumbered}

In real bioinformatics work, data files are rarely perfect. They may have **missing headers**, **non-standard delimiters**, **comment lines**, **European decimal marks**, **embedded special characters**, or **various missing-value representations**. The following exercises train you to handle these messy files. 在真实生物信息学工作中，数据文件很少是完美的。它们可能有缺失的列名、非标准分隔符、注释行、欧洲小数点、嵌入的特殊字符或各种缺失值表示。以下练习训练你处理这些混乱文件。

All data files are located in `../Exercises and homework/data/`. 所有数据文件位于 `../Exercises and homework/data/`。

**Important**: For each file, provide **two methods**: (1) using base R functions (`read.table` / `read.csv`), and (2) using `readr` package functions (`read_delim` / `read_csv` / `read_csv2`). Pay attention to the data characteristics and read each column into the correct data type. 对于每个文件，提供两种方法读取。

## 5.6 Table0.txt — no header, tab-separated

This file contains 7 rows of person data (Name, Age, Height, Weight, Sex) but **has no column header**. Delimiter is **tab**. 该文件无列名，制表符分隔。

```{r}
## Method 1: Base R — read.table()
```

```{r}
## Method 2: readr — read_delim()
```

## 5.7 Table1.txt — with header, tab-separated

Same data as Table0 but **with a header row**. Still tab-delimited. 与 Table0 数据相同，但有列名行。

```{r}
## Method 1: Base R
```

```{r}
## Method 2: readr
```

## 5.8 Table2.txt — comment line + `/` wrappers

This file has **two header lines** (a comment "Table 2: ..." then the real header). Names are wrapped with `/` (e.g., `/Alex/`), and Sex is wrapped with `/` (e.g., `/F/`). You need to **skip the comment line** and **strip the `/` characters**. 该文件有两行表头（一行注释 + 一行列名），姓名和性别被 `/` 包裹，需要跳过注释行并去除 `/` 字符。

```{r}
## Method 1: Base R — use skip=, then clean with gsub()
```

```{r}
## Method 2: readr — use skip=, comment=, then mutate with str_remove_all()
library(stringr)
```

**Question**: What `comment` argument value did you use for readr? Why? 你用了什么 `comment` 参数值？为什么？

**Answer**:

## 5.9 Table3.txt — various missing value representations

This file has **two header lines** like Table2, but **no `/` wrapping**. The challenge: missing values are represented as `NA` (Age for Lilly), `--` (Age for Mark), `*` (Height for Martha), and `**` (Weight for Lucas). You must convert all of these to `NA` in R. 该文件包含多种缺失值表示方式，需要将它们都转换为 R 的 NA。

```{r}
## Method 1: Base R — use na.strings argument
```

```{r}
## Method 2: readr — use na argument
```

**Question**: What are the different `na` representations you specified? List them all. 你指定了哪些缺失值表示方式？请全部列出。

**Answer**:

## 5.10 Table4.txt — European decimal marks (`,` as decimal)

This file has **two header lines** and missing values (like Table3), but with **European decimal format**: Height values like `1,77` mean `1.77` (comma as decimal separator). You must read Height as `col_double()`. 该文件使用欧洲小数点格式，逗号表示小数点，Height 列需读取为数值型。

```{r}
## Method 1: Base R — use dec="," or read.csv2() approach
## Hint: read the file as character first, then convert comma to dot and coerce to numeric
```

```{r}
## Method 2: readr — use locale = locale(decimal_mark = ",")
```

**Question**: How does `locale(decimal_mark = ",")` change the parsing behavior? `locale(decimal_mark = ",")` 如何改变解析行为？

**Answer**:

## 5.11 Table5.txt — semicolon-delimited + European decimals

This file uses **`;` as delimiter** (semicolon-separated), with **European decimal marks** and **various missing values**. No comment lines, but missing values include `--`, `*`, `**`, and empty fields. 该文件以分号分隔，使用欧洲小数点，包含多种缺失值。

```{r}
## Method 1: Base R — use read.csv2() or read.table(sep=";")
```

```{r}
## Method 2: readr — use read_csv2() or read_delim(delim=";")
```

## 5.12 Table6.txt — repeated data with annotations

This file has **108 lines**: 2 header lines + 105 data lines (7 people × 15 repetitions). Some data lines have annotations like `@Boss` and `@Secretary` appended to the Sex column. You need to:
- Skip the 2 comment/header lines
- **Remove duplicate rows** after reading
- **Strip the `@...` annotations** from the Sex column

该文件包含重复数据和 `@Boss`/`@Secretary` 等注释，需要去重并清理 Sex 列。

```{r}
## Method 1: Base R — use skip=, then unique() and gsub()
```

```{r}
## Method 2: readr — use skip=, then distinct() and str_remove()
library(dplyr)
library(stringr)
```

**Question**: How many unique rows remain after deduplication? 去重后剩下多少唯一行？

**Answer**:

## 5.13 states1.csv and states2.csv — compare two formats

These two files contain the same US states data but in **different formats**:

- **states1.csv**: Standard CSV (comma-separated), with **European decimals** (`2,1` means 2.1), **quoted state names**
- **states2.csv**: Semicolon-separated (`;`), **quoted fields**, **European decimals**

Read both files and **verify they produce identical data frames**. 读取两个文件，验证它们产生相同的数据框。

```{r}
## Read states1.csv — Method 1: Base R
## Hint: use read.csv() or read.table() with appropriate arguments
```

```{r}
## Read states1.csv — Method 2: readr
```

```{r}
## Read states2.csv — Method 1: Base R
```

```{r}
## Read states2.csv — Method 2: readr
```

```{r}
## Verify both produce identical data
## Use identical() or all.equal() to compare
```

**Question**: What were the key differences in arguments between reading states1.csv and states2.csv? 读取 states1.csv 和 states2.csv 时，参数的关键区别是什么？

**Answer**:

**Question**: After this exercise, which approach (base R vs readr) do you find easier to use for messy real-world files? Why? 完成本练习后，对于混乱的真实文件，你觉得哪种方式（base R 还是 readr）更好用？为什么？

**Answer**:

---

# Exercise 6: Dates with lubridate 日期处理 {.unnumbered}

## 6.1 Parse dates in different formats

Parse the following date strings using `lubridate` functions. For each, display the parsed result and extract the year and month. 用 lubridate 解析日期字符串，提取年月。

```{r}
library(lubridate)

# (a) "2024-03-15"
# (b) "March 15, 2024"
# (c) "15-03-2024"
# (d) "2024-03-15 14:30:00" (date-time)
```

## 6.2 Date arithmetic

Calculate: 日期算术运算。

```{r}
# (a) What is the date 90 days from today?
# (b) How many days between "2024-01-01" and "2024-12-31"?
# (c) What day of the week was "2020-01-23"? (Use wday with label = TRUE)
```

## 6.3 Sample collection timeline

A clinical study collected samples on these dates: `"2024-01-15"`, `"2024-02-20"`, `"2024-03-10"`, `"2024-04-05"`, `"2024-05-12"`. 计算采样时间间隔。

```{r}
# (a) Parse all dates into a vector
# (b) Calculate the interval (in days) between the first and last collection
# (c) Calculate the interval between each consecutive pair of dates
```

---

# Exercise 7: Putting It All Together 综合练习 {.unnumbered}

## 7.1 Build a gene study data table

Create a tibble that combines all concepts from this talk:

- Use `tibble()` to create a table with columns: `gene` (factor), `log2FC` (numeric), `pvalue` (numeric), and `collection_date` (date)
- The `gene` factor should have levels: `"BRCA1"`, `"TP53"`, `"EGFR"`, `"MYC"`
- Use `ymd()` for the dates
- After creating, filter to show only rows where `pvalue < 0.05`

综合运用本讲所有概念创建数据表。

```{r}
library(tibble)
library(lubridate)
library(forcats)

## Write your code here — create the tibble with all requirements
```

```{r}
# Filter for pvalue < 0.05
```

## 7.2 Write results and read back

```{r}
# (a) Write the filtered results (pvalue < 0.05) to "significant_genes.csv"
# (b) Read the file back to verify
```

---

# Exercise 8: Use CodeBuddy to Help 使用 AI 辅助 {.unnumbered}

## 8.1 Debug with AI

If you encountered any errors in the exercises above, paste the error message to CodeBuddy and ask for help. Record the prompt and summary below. 如遇到错误，用 CodeBuddy 调试。

**My debug prompt**:

```
```

**CodeBuddy's response summary**:

## 8.2 Explain a concept with AI

Ask CodeBuddy: "Explain the difference between a factor and a character vector in R. When should I use each?"

**My prompt**:

```
```

**CodeBuddy's response summary**:

## 8.3 Generate extra practice

Ask CodeBuddy: "Give me 3 additional exercises about R lists, with answers. Focus on nested list extraction."

**My prompt**:

```
```

**CodeBuddy's response summary**:

---

*Homework ends here. Make sure all code chunks run without errors, then Render to PDF and submit. 作业结束。请确认所有代码块运行无误，Render 生成 PDF 后提交。*
