---
title: "Talk 05 — Homework"
subtitle: "Data Visualization I — Base R & ggplot2 Fundamentals"
author: "Your Name — Your Student ID"
date: today
CJKmainfont: "Hei"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    pdf-engine: xelatex
  html:
    df-print: paged
---

## Instructions 说明

1.  All exercises require both **base R graphics** and **ggplot2** versions unless stated otherwise. 除非特别说明，所有练习都需要同时提供 base R 和 ggplot2 两个版本。
2.  Run each code chunk to verify results. 运行每个代码块以验证结果。
3.  Submit the compiled PDF as `姓名-学号-talk05-homework.pdf`. 提交编译后的 PDF。

## User Verification 用户验证

```{r}
Sys.info()[["user"]]
Sys.getenv("HOME")
```

```{r}
library(tidyverse)
```

---

# Exercise 1: Basic Plots with swiss Data 基础作图 {.unnumbered}

## 1.1 Histogram of Catholic

Use **both** base R `hist()` and ggplot2 `geom_histogram()` to show the distribution of `Catholic` from the `swiss` dataset.

```{r}
## Base R version

```

```{r}
## ggplot2 version

```

## 1.2 Scatter Plot

Plot `Education` vs. `Fertility` from `swiss` with proper axis labels and a title. Provide **both** versions.

```{r}
## Base R version — use plot()

```

```{r}
## ggplot2 version — use geom_point()

```

**Question**: Which approach do you find more intuitive for creating a scatter plot? Why? 哪种方法创建散点图更直观？为什么？

**Answer**:

---

# Exercise 2: iris Data 鸢尾花数据 {.unnumbered}

## 2.1 Scatter Plot by Species

Plot `Sepal.Length` vs. `Petal.Length`, colored by `Species`, with a legend. Provide **both** versions.

```{r}
## Base R version — use plot() with col and legend()

```

```{r}
## ggplot2 version

```

## 2.2 Box Plot by Species

Show the distribution of `Sepal.Length` across `Species` using box plots. Provide **both** versions.

```{r}
## Base R version — use boxplot()

```

```{r}
## ggplot2 version — use geom_boxplot()

```

---

# Exercise 3: Multiple Geoms {.unnumbered}

## 3.1 Scatter + Smooth

Using `mtcars`, plot `wt` vs. `mpg`. Add both points (`geom_point`) and a LOESS smooth line (`geom_smooth`). Color points by `cyl` (as factor).

```{r}
## ggplot2 version

```

## 3.2 Describe the Trend

**Question**: What is the relationship between car weight and fuel efficiency? What does the smooth line tell you? 车重与油耗的关系是什么？平滑线说明了什么？

**Answer**:

---

# Exercise 4: Scales 标尺 {.unnumbered}

## 4.1 Color Scales

Using `mpg` dataset, create a scatter plot of `displ` vs. `hwy`:

1. Color points by `class` using `scale_color_brewer(palette = "Set1")`
2. Create the same plot but use `scale_color_manual()` with your own colors
3. Which palette type (sequential, qualitative, diverging) is appropriate here? Why?

```{r}
## 1. scale_color_brewer

```

```{r}
## 2. scale_color_manual

```

**Answer to Q3**:

## 4.2 Gradient Color

Using `mpg` dataset, plot `displ` vs. `hwy` and color points by `cty` (continuous variable) using `scale_color_gradient()`.

```{r}
## Your code here

```

---

# Exercise 5: Coordinate Systems 坐标系 {.unnumbered}

## 5.1 coord_flip

Using `mpg`, create a box plot of `hwy` by `class` (as factor). Then flip the coordinates. When is `coord_flip` useful?

```{r}
## Your code here

```

**Answer**: 

## 5.2 coord_polar

Using `mtcars`, create a bar chart of `cyl` counts, then convert it to a pie chart with `coord_polar()`. Hint: start with `geom_bar()`.

```{r}
## Your code here

```

---

# Exercise 6: Facets 分面 {.unnumbered}

## 6.1 Facet by cyl

Plot `wt` vs. `mpg` from `mtcars`, faceted by `cyl` using `facet_grid()`. Add `geom_point()` + `geom_smooth(method = "lm")`.

```{r}
## Your code here

```

## 6.2 Facet airquality

Use `airquality` data. Plot `Wind` vs. `Temp` faceted by `Month` using `facet_wrap()` (2 rows × 3 columns). Add `geom_point()` + `geom_smooth()`.

```{r}
## Your code here

```

## 6.3 Free Scales

Re-do 6.2 but use `scales = "free"` in `facet_wrap()`. What changed? When would you use free scales?

```{r}
## Your code here

```

**Answer**:

---

# Exercise 7: AI-Assisted Practice AI 辅助练习 {.unnumbered}

## 7.1 Debug with AI

Take any plot from this homework that gave you trouble. Ask CodeBuddy/your AI tool to:
1. Explain what went wrong
2. Suggest the fix
3. Show the corrected code

Paste the AI's response below.

**AI Response**:

```

```

## 7.2 Generate New Exercises

Ask your AI tool: *"Give me 3 ggplot2 exercises using the `diamonds` dataset that test scales, facets, and coordinate systems."*

**AI Response**:

```

```

## 7.3 Explain a Concept

Ask your AI tool: *"Explain the difference between `colour` and `fill` in ggplot2 with examples. Also explain when to use `facet_grid()` vs. `facet_wrap()`."*

**AI Response**:

```

```
