---
title: "Talk 06 — Homework"
subtitle: "Data Visualization II — Advanced ggplot2, Themes & Saving"
author: "Your Name — Your Student ID"
date: today
CJKmainfont: "Hei"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    pdf-engine: xelatex
  html:
    df-print: paged
---

## Instructions 说明

1.  Run each code chunk to verify results. 运行每个代码块以验证结果。
2.  Submit the compiled PDF as `姓名-学号-talk06-homework.pdf`.
3.  All plots should have proper labels and legends. 所有图都要有合适的标签和图例。

## User Verification 用户验证

```{r}
Sys.info()[["user"]]
Sys.getenv("HOME")
```

```{r}
library(tidyverse)
library(cowplot)
library(gridExtra)
```

---

# Exercise 1: Multi-panel with cowplot 多图组合 {.unnumbered}

## 1.1 Prepare Four Plots

Create the following four plots and save them:

- **p1**: `starwars` — box plot of `height` by `gender` (filter NA heights, keep top 3 genders by count)
- **p2**: `iris` — scatter of `Sepal.Length` vs. `Sepal.Width`, colored by `Species`, with `geom_smooth()`
- **p3**: `mtcars` — scatter of `wt` vs. `mpg`, colored by `cyl` (as factor)
- **p4**: `ToothGrowth` — violin plot of `len` by `dose` (as factor), filled by `supp`

```{r}
## Your code here — save as p1, p2, p3, p4

```

## 1.2 plot_grid

Combine all four plots using `cowplot::plot_grid()` in a 2×2 layout with labels A, B, C, D.

```{r}
## Your code here

```

## 1.3 ggdraw Free Layout

Using `ggdraw()` + `draw_plot()`, arrange p1 and p2 in the top row (each 50% width), and p3 spanning the full bottom row. Add labels.

```{r}
## Your code here

```

---

# Exercise 2: gridExtra Layouts 矩阵布局 {.unnumbered}

## 2.1 grid.arrange

Create three plots from `diamonds`:
- **g1**: bar plot of `cut`
- **g2**: box plot of `price` by `cut`
- **g3**: histogram of `carat`

Arrange them using `grid.arrange()` with a layout matrix that puts g1 on the left (full height) and g2/g3 stacked on the right.

```{r}
## Your code here

```

---

# Exercise 3: Equations in Plots 图中公式 {.unnumbered}

## 3.1 Regression Equation

Using the `swiss` dataset, fit a linear model: `Fertility ~ Education`. Create a ggplot2 scatter plot with:

1. Points and a smooth line
2. The regression equation in the form: *y* = *a* + *b*·*x*
3. *R*² and *p*-value
4. Use `substitute()` or `bquote()` to add the equation as text

```{r}
## Your code here

```

## 3.2 Greek Letters

Create a plot that displays all 24 Greek letters (both lowercase and uppercase) using `geom_text()` with `parse = TRUE`. Arrange them in a grid.

```{r}
## Your code here

```

---

# Exercise 4: Themes & Customization 主题与自定义 {.unnumbered}

## 4.1 Theme Comparison

Using `mpg` dataset, create a box plot of `hwy` by `class`. Create 4 versions, each with a different built-in theme (`theme_bw`, `theme_minimal`, `theme_classic`, `theme_light`). Arrange them in a 2×2 layout.

```{r}
## Your code here

```

**Question**: Which theme would you use for a publication figure? Why? 哪个主题适合发表？为什么？

**Answer**:

## 4.2 Custom theme()

Take the `theme_bw` version from 4.1 and customize it:
1. Move legend to the bottom
2. Rotate x-axis labels 45 degrees
3. Increase base font size to 14
4. Remove minor grid lines

```{r}
## Your code here

```

## 4.3 labs() Mastery

Using the `iris` dataset, create a scatter plot of `Sepal.Length` vs. `Petal.Length`, colored and shaped by `Species`. Use `labs()` to set:
- Custom x and y labels
- Legend titles for color and shape (merged into one)
- A title, subtitle, and caption

```{r}
## Your code here

```

---

# Exercise 5: Saving Plots 保存图形 {.unnumbered}

## 5.1 ggsave Practice

Save the plot from Exercise 4.3 as both PDF (vector) and PNG (raster, 300 dpi). Use `ggsave()`.

```{r}
## Your code here

```

**Question**: When would you use PDF vs. PNG for saving figures? PDF和PNG各适用于什么场景？

**Answer**:

## 5.2 Interactive Plot

Install `plotly` (if not already) and convert the plot from Exercise 4.3 to an interactive version using `ggplotly()`. What additional features does the interactive version provide?

```{r}
## Your code here

```

**Answer**:

---

# Exercise 6: AI-Assisted Practice AI 辅助练习 {.unnumbered}

## 6.1 Multi-panel Help

Ask your AI tool: *"How do I create a publication-quality multi-panel figure in R with shared legends and consistent styling? Show me with example code."*

**AI Response**:

```

```

## 6.2 Theme Customization

Ask your AI tool: *"Show me how to create a custom ggplot2 theme that mimics the style of Nature or Science journals."*

**AI Response**:

```

```

## 6.3 Debug

If any exercise above gave you errors, paste the error message here and ask your AI tool for help. Record the solution.

**Error**:

```

```

**AI Solution**:

```

```
